---
title: "Map optimal genotypes by climate"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    number_sections: true
    highlight: rstudio
---

Using reaction norms modeled in the 'transfer_function_multiyear_linear_mixed_effects_model.Rmd' script, identify the genotype having highest growth (taking into account mortality probability) across the sampled hybrid zone under historic and future values of MCMT. Maps show where the 'optimal' genotype/species ancestry could shift spatially under future temperatures, resulting in spatial shifts in the location of the hybrid zone.

# Setup

## Load packages and data
```{r maps_setup}

library(sf) # mapping
library(terra) # mapping
library(RColorBrewer) # colors
library(circlize) # colorRamp2

# predictions of growth and mortality across climate for each genotype
load('results/model_prediction/predictedGrowth_acrossClimates_byGenotype_garden_MCMT_2021-2022.Rdata')
# rename
preds <- predictions
rm(predictions)

# make genotype info its own df
geno.info <- preds$genotypes
#cleanup
geno.info$genotype <- rownames(geno.info)
rownames(geno.info) <- paste('genotype_', rownames(geno.info), sep = '')


# garden data
load('data/clean/garden_climates_average1961-1990_yearly2020-2023.Rdata')
gards <- clim

# provenance data
# extract from garden climate and phenotype origin data - named 'dat'
load('data/clean/mini_garden_phenotypic_and_climate_data_2021-2023.Rdata')
# get first instance of genotype, extract genotype info
provs <- dat[! duplicated(dat$Genotype), c("Genotype", "Pb", "Pt", 
"transect", "provenance_latitude", "provenance_longitude", "provenance_elevation_m")]
# remove NA genotype
provs <- provs[! is.na(provs$Genotype),]
# remove genotypes without genetic data
provs <- provs[! is.na(provs$Pt),]

# change rownames to genotypes
rownames(provs) <- provs$Genotype

sessionInfo()

```

```{r markdown_settings}

knitr::opts_chunk$set(fig.width = 12, 
                      fig.height = 10)

```

## Setup genotype info

```{r genotype_info}

# get additional genotype info

# get first instances of each genotype
genos <- dat[match(geno.info$genotype, dat$Genotype),]

# check for match
cbind(as.character(genos$Genotype), geno.info$genotype)

# add to dataframe
geno.info$transect <- genos$transect
geno.info$interspecific_heterozygosity <- genos$interspecific_heterozygosity
geno.info$heterozygosity <- genos$heterozygosity
geno.info$hybrid_index <- genos$hybrid_index
geno.info$plastid_ID <- genos$plastid_ID
geno.info$Pt <- genos$Pt
geno.info$color_Pt <- genos$color_Pt
geno.info$color_k3 <- genos$color_k3

str(geno.info)

```


## Merge genotype responses

Merge each genotype's response to MCMT into one dataframe for comparison across genotypes

```{r merge_genotypes_predictions}

# get the overall predictions by genotype and join into one dataframe to identify optimal genotype for each climate

# get first genotype
pred.overall <- preds$overall[[1]]

# rename column
colnames(pred.overall)[2] <- names(preds$overall)[1]


# loop through other genotypes and add them
for(n in 2:length(preds$overall)){
  
  df <- preds$overall[[n]]
  
  pred.overall[,names(preds$overall)[n]] <- df[,2]
  
}

head(pred.overall)

# which genotype has highest predicted success for each value of MCMT (row)?

# just get genotypes
tmp <- pred.overall[,-1]


pred.overall$best_genotype <- sapply(1:nrow(tmp), function(x) names(which.max(tmp[x,])))

# get genotype colors
pred.overall$best_genotype_col_Pt <- geno.info[pred.overall$best_genotype, 'color_Pt']
pred.overall$best_genotype_col_k3 <- geno.info[pred.overall$best_genotype, 'color_k3']

pred.overall$best_genotype <- gsub('genotype_', '', pred.overall$best_genotype)

# quick plot to check
plot(pred.overall$garden_clim, col = pred.overall$best_genotype_col_Pt, pch = 16)
plot(pred.overall$garden_clim, col = pred.overall$best_genotype_col_k3, pch = 16)

# list 'best' genotype
cbind(pred.overall$garden_clim, pred.overall$best_genotype)

# save
# save(pred.overall, file = 'results/model_prediction/genotypePredictedHeights_byGardenMCMT.Rdata')
# write.csv(pred.overall, file = 'results/model_prediction/genotypePredictedHeights_byGardenMCMT.csv', row.names = F)

```

## Load climate rasters

Load future and historic MCMT raster files

ClimateNA raster files used here are available from DataBasin:

https://adaptwest.databasin.org/pages/adaptwest-climatena/

Species range shapefiles are from Little 1971 and are available from DataBasin:

P. balsamifera: https://databasin.org/datasets/91380e091ca048359a66fc65962ed210/

P. trichocarpa: https://databasin.org/datasets/84e47784fe2a463c8b292007fb43f2cd/

```{r get_climate}

# which climate variable is being used? 
clim_var <- 'MCMT'
# calculate 2-year average, used in model
dat$garden_MCMT_2020_2021_avg <- (dat$garden_MCMT_2020 + dat$garden_MCMT_2021)/2
gard_clim_colname <- 'garden_MCMT_2020_2021_avg'

# get historic and future climate

hist <- rast(paste('data/climate/climateNA/Normal_1961_1990/Normal_1961_1990_bioclim/Normal_1961_1990_', clim_var, '.tif', sep = ''))
crs(hist, proj = T)

hist91 <- rast(paste('data/climate/climateNA/Normal_1991_2020/Normal_1991_2020_', clim_var, '.tif', sep = ''))
crs(hist, proj = T)

fut8 <- rast(paste('data/climate/climateNA/future/ensemble_8GCMs_ssp245_2041_2070_bioclim/ensemble_8GCMs_ssp245_2041_2070_', clim_var, '.tif', sep = ''))

fut13 <- rast(paste('data/climate/climateNA/future/ensemble_13GCMs_ssp245_2041_2070_bioclim/ensemble_13GCMs_ssp245_2041_2070_', clim_var, '.tif', sep = ''))

################ convert coords

# convert coords to crs of climateNA
crs.cna <- crs(hist, proj = T)

# provenance coordinates
coords.prov <- st_as_sf(provs[,c("provenance_longitude", "provenance_latitude")], coords = c(1,2), crs = st_crs(4326))
coords.prov.cna <- st_transform(coords.prov$geometry, crs = crs.cna)

# garden coordinates
coords.gards <- st_as_sf(gards[,c("Longitude", "Latitude")], coords = c(1,2), crs = st_crs(4326))
coords.gards.cna <- st_transform(coords.gards$geometry, crs = crs.cna)

# load shapefile of state/province borders
borders <- read_sf('data/shapefiles/NorthAmerica_PoliticalBoundaries_Shapefile/NA_PoliticalDivisions/data/bound_p/boundaries_p_2021_v3.shp')
borders <- st_transform(borders, crs = crs.cna)
# simplify to 1 km
borders <- st_simplify(borders, dTolerance = 1000)

# species rangemaps
# shapefile
balsam <- st_read('data/shapefiles/Pbal_shapefile/data/commondata/data0/popubals.shp')
tricho <- st_read('data/shapefiles/Ptri_shapefile/data/commondata/data0/poputric.shp')

# convert CRS
balsam <- st_transform(balsam, crs.cna)
tricho <- st_transform(tricho, crs.cna)

# test plot
plot(hist)

plot(balsam$geometry, add = T, border = 'navy')
plot(tricho$geometry, add = T, border = 'grey50')
plot(borders$geometry, add = T)


```

# Mapping (color by species ancestry)

## Functions

```{r map_best_geno_functions}

# function for plotting
map_best_geno <- function(raster, raster_name, save = F, ...){
  
  if(save == T){
    png(file = paste('results/model_prediction/best_genotype_2yearModel_mapped_by_', clim_var, '_', raster_name, '.png', sep = ''),
   height = 8, width = 8, res = 300, units = 'in')
  }
  
  
  plot(raster,
       pax = list(side=NA),
       colNA = 'white',
       ...)
  
  plot(borders$geometry, add = T, lwd = 0.5)
  plot(balsam$geometry, add = T, border = 'navy')
  plot(tricho$geometry, add = T, border = 'darkgreen')
  plot(coords.prov.cna, add = T, lwd = 2, bg = geno.info$color_Pt, pch = 21, cex = 1.5)
  plot(coords.gards.cna, add = T, pch = 17, cex = 2)
  
  # legend(-3e6, -1.8e6, 
  #        pch = c(22,22,21,24),
  #        pt.cex = 2,
  #        xjust = 0.5,
  #        pt.bg =  c('darkgreen', 'navy', 'white','white'), 
  #        legend = c(substitute(paste(italic('P. trichocarpa'), ' range')), 
  #                   substitute(paste(italic('P. balsamifera'), ' range')),
  #                   'Collection Site',
  #                   'Common Garden Site'))
  
    legend(-3e6, -1.8e6, 
         pch = c(22,22,21,24),
         pt.cex = 2,
         xjust = 0.5,
         col =  c('darkgreen', 'navy', 'black','black'), 
         pt.lwd = 2,
         pt.bg = c('white', 'white', 'white', 'white'),
         legend = c(substitute(paste(italic('P. trichocarpa'), ' range')), 
                    substitute(paste(italic('P. balsamifera'), ' range')),
                    'Collection Site',
                    'Common Garden Site'))
  
  
  
  # add inset legend for species ancestry gradient
  # https://stackoverflow.com/questions/13355176/gradient-legend-in-base
  
  # get breaks for legend labels from color palette
  pal_range <- range(attr(colf.pt, 'breaks'))
  pal_min <- pal_range[1]
  pal_max <- pal_range[2]
  
  legend_image <- as.raster(matrix(colf.pt(seq(pal_max, pal_min, length = 100)), ncol=1))
  
  # look at NDC coords
  grconvertX(seq(-4e6, 3e6, 1000000), from = 'user', to = 'ndc')
  grconvertY(seq(-3e6, 4e6, 1000000), from = 'user', to = 'ndc')
  
  figSet <- c(0.1, 0.4, 0.15, 0.5)
  op <- par(  ## set and store par
    fig=figSet,    ## set figure region, 
    mar=c(1, 1, 1, 9.5),                                  ## set margins
    new=TRUE)                                ## set new for overplot w/ next plot
  
  plot(0,0, type='n', axes=F, xlab='', ylab='')  ## ini plot2
  rasterImage(legend_image, 0, 0, 1, 1)                       ## the gradient
  lbsq <- seq.int(0, 1, l=5)                                  ## seq. for labels
  axis(4, at=lbsq, pos=1, labels=F, col=0, col.ticks=1, tck=-.1)  ## axis ticks
  mtext(lbsq, 4, 0.3, at=lbsq, las=2, cex=.8)                    ## tick labels
  
  mtext(expression(atop('Best-performing\ngenotype', italic('(% P. trichocarpa)'))), side=3, line=0.2, cex=1, adj=.1)          ## legend title
  
  par(op)  ## reset par
  
  
  if(save == T){
    dev.off()
  }
}

######################################
# inset zoomed to collection sites


# function
map_best_geno_crop <- function(raster, raster_name, save = F, ...){
  
  if(save == T){
    png(file = paste('results/model_prediction/best_genotype_2yearModel_mapped_by_', clim_var, '_inset_', raster_name, '.png', sep = ''),
   height = 8, width = 6, res = 300, units = 'in')
  }
  
  
  plot(raster,
       pax = list(side=NA),
       colNA = 'white',
       ...)
  
  plot(borders$geometry, add = T, lwd = 0.5)
  plot(balsam$geometry, add = T, border = 'navy')
  plot(tricho$geometry, add = T, border = 'darkgreen')
  plot(coords.prov.cna, add = T, lwd = 2, bg = geno.info$color_Pt, pch = 21, cex = 2)
  
  # legend(-3e6, -1.8e6, 
  #        pch = c(22,22,21,24),
  #        pt.cex = 2,
  #        xjust = 0.5,
  #        pt.bg =  c('darkgreen', 'navy', 'white','white'), 
  #        legend = c(substitute(paste(italic('P. trichocarpa'), ' range')), 
  #                   substitute(paste(italic('P. balsamifera'), ' range')),
  #                   'Collection Site',
  #                   'Common Garden Site'))
  
  if(save == T){
    dev.off()
  }
  
}
  
```


## Plot

```{r plot}

# color scale - used in legend
colf.pt <- colorRamp2(c(0,0.5,1), colors = c('dodgerblue2', 'grey20', 'darkolivegreen2'))

# assign color of 'optimal' genotype to each temperature bin using coltab 
cols <- cbind(pred.overall$garden_clim, pred.overall$best_genotype_col_Pt)
head(cols)

# set raster colors
coltab(hist) <- cols
coltab(hist91) <- cols
coltab(fut8) <- cols
coltab(fut13) <- cols

# make values outside prediction range (common gardens) NA
# get the range of actual garden climates that were tested
pred_clim <- c(dat$garden_MCMT_2021, dat$garden_MCMT_2022)

msk <- ifel(hist > max(pred_clim) | hist < min(pred_clim), NA, 1)
hist.mask <- mask(hist, msk)
plot(hist.mask, colNA = 'grey')

hist91.mask <- mask(hist91, msk)
plot(hist91.mask, colNA = 'grey')

# future climate

# make values outside predict range NA
msk <- ifel(fut13 > max(pred_clim) | fut13 < min(pred_clim), NA, 1)
fut13.mask <- mask(fut13, msk)
plot(fut13.mask, colNA = 'grey')

msk <- ifel(fut8 > max(pred_clim) | fut8 < min(pred_clim), NA, 1)
fut8.mask <- mask(fut8, msk)
plot(fut8.mask, colNA = 'grey')

# are we saving plots?
save = FALSE

# Maps!

# plot full range
map_best_geno(hist.mask, raster_name = 'historic', main = '1961-1990', save = save)
map_best_geno(hist91.mask, raster_name = '1991-2020', main = '1991-2020', save = save)
map_best_geno(fut13.mask, raster_name = '2041-2070_13GCMs', main = '2041-2070 (13 GCM ensemble)', save = save)
map_best_geno(fut8.mask, raster_name = '2041-2070_8GCMs', main = '2041-2070 (8 GCM ensemble)', save = save)


# crop rasters, then plot
extent <- ext(-26e5, -5e5, 45e4, 3e6)
hist.crop <- crop(hist.mask, extent)
hist91.crop <- crop(hist91.mask, extent)
fut8.crop <- crop(fut8.mask, extent)
fut13.crop <- crop(fut13.mask, extent)

map_best_geno_crop(hist.crop, main = '1961-1990', save = save, raster_name = 'historic')
map_best_geno_crop(hist91.crop, raster_name = '1991-2020', main = '1991-2020', save = save)
map_best_geno_crop(fut8.crop, main = '2041-2070 (8 GCM ensemble)', save = save, raster_name = '2041-2070_8GCMs')
map_best_geno_crop(fut13.crop, main = '2041-2070 (13 GCM ensemble)', save = save, raster_name = '2041-2070_13GCMs')


# crop to western interior region

extent <- ext(-15e5, 1e5, -10e5, 1e6)
hist.crop <- crop(hist.mask, extent)
hist91.crop <- crop(hist91.mask, extent)
fut8.crop <- crop(fut8.mask, extent)
fut13.crop <- crop(fut13.mask, extent)

map_best_geno_crop(hist.crop, main = '1961-1990', save = save, raster_name = 'west_historic')
map_best_geno_crop(hist91.crop, raster_name = 'west_1991-2020', main = '1991-2020', save = save)
map_best_geno_crop(fut8.crop, main = '2041-2070 (8 GCM ensemble)', save = save, raster_name = 'west_2041-2070_8GCMs')
map_best_geno_crop(fut13.crop, main = '2041-2070 (13 GCM ensemble)', save = save, raster_name = 'west_2041-2070_13GCMs')


```

# Mapping (color by K=3)

Here, color by ancestry based on admixture with K=3 (separating tricho into a coastal and interior lineage). Colors for each genotype are calculated as RGB values with blue = balsam, red = interior tricho, and green = coastal tricho.

Did not use these in the manuscript because the tricho/balsam pattern seems to be stronger, and genetic PC2, which separates coastal and interior tricho, was not significant in the model of growth in the common gardens.

## Functions

```{r map_best_color_k3_functions}

# function for plotting
# comapred to function above, this just removes the ancestry scale

map_best_geno_k3 <- function(raster, raster_name, save = F, ...){
  
  if(save == T){
    png(file = paste('results/model_prediction/best_genotype_2yearModel_colorK3_mapped_by_', clim_var, '_', raster_name, '.png', sep = ''),
   height = 8, width = 8, res = 300, units = 'in')
  }
  
  
  plot(raster,
       pax = list(side=NA),
       colNA = 'white',
       ...)
  
  plot(borders$geometry, add = T, lwd = 0.5)
  plot(balsam$geometry, add = T, border = 'navy')
  plot(tricho$geometry, add = T, border = 'darkgreen')
  plot(coords.prov.cna, add = T, lwd = 2, bg = geno.info$color_k3, pch = 21, cex = 1.5)
  plot(coords.gards.cna, add = T, pch = 17, cex = 2)
  
  legend(-3e6, -1.8e6, 
         pch = c(22,22,21,24),
         pt.cex = 2,
         xjust = 0.5,
         pt.bg =  c('darkgreen', 'navy', 'white','white'), 
         legend = c(substitute(paste(italic('P. trichocarpa'), ' range')), 
                    substitute(paste(italic('P. balsamifera'), ' range')),
                    'Collection Site',
                    'Common Garden Site'))
  
  
  if(save == T){
    dev.off()
  }
}


# inset / zoomed

map_best_geno_crop_k3 <- function(raster, raster_name, save = F, ...){
  
  if(save == T){
    png(file = paste('results/model_prediction/best_genotype_2yrModel_colorK3_mapped_by_', clim_var, '_inset_', raster_name, '.png', sep = ''),
   height = 8, width = 6, res = 300, units = 'in')
  }
  
  
  plot(raster,
       pax = list(side=NA),
       colNA = 'white',
       ...)
  
  plot(borders$geometry, add = T, lwd = 0.5)
  plot(balsam$geometry, add = T, border = 'navy')
  plot(tricho$geometry, add = T, border = 'darkgreen')
  plot(coords.prov.cna, add = T, lwd = 2, bg = geno.info$color_k3, pch = 21, cex = 2)
  
  
  if(save == T){
    dev.off()
  }
}
```


## Plot

```{r plot_k3}

# color by K=3
cols <- cbind(pred.overall$garden_clim, pred.overall$best_genotype_col_k3)

# set raster colors
coltab(hist.mask) <- cols
coltab(hist91.mask) <- cols
coltab(fut8.mask) <- cols
coltab(fut13.mask) <- cols

# are we saving plots?
save = FALSE

# map

map_best_geno_k3(hist.mask, raster_name = 'historic', main = 'Historic (1961-1990)', save = save)
map_best_geno_k3(hist91.mask, raster_name = '1991-2020', main = '1991-2020', save = save)
map_best_geno_k3(fut8.mask, raster_name = '2041-2070_8GCMs', main = '2041-2070 (8 GCM ensemble)', save = save)
map_best_geno_k3(fut13.mask, raster_name = '2041-2070_13GCMs', main = '2041-2070 (13 GCM ensemble)', save = save)

# crop rasters, then plot
extent <- ext(-26e5, -5e5, 45e4, 3e6)
hist.crop <- crop(hist.mask, extent)
hist91.crop <- crop(hist91.mask, extent)
fut8.crop <- crop(fut8.mask, extent)
fut13.crop <- crop(fut13.mask, extent)

map_best_geno_crop_k3(hist.crop, main = '1961-1990', save = save, raster_name = 'historic')
map_best_geno_crop_k3(hist91.crop, raster_name = '1991-2020', main = '1991-2020', save = save)
map_best_geno_crop_k3(fut8.crop, main = '2041-2070 (8 GCM ensemble)', save = save, raster_name = '2041-2070_8GCMs')
map_best_geno_crop_k3(fut13.crop, main = '2041-2070 (13 GCM ensemble)', save = save, raster_name = '2041-2070_13GCMs')

# crop to western interior region

extent <- ext(-15e5, 1e5, -10e5, 1e6)
hist.crop <- crop(hist.mask, extent)
hist91.crop <- crop(hist91.mask, extent)
fut8.crop <- crop(fut8.mask, extent)
fut13.crop <- crop(fut13.mask, extent)

map_best_geno_crop_k3(hist.crop, main = '1961-1990', save = save, raster_name = 'west_historic')
map_best_geno_crop_k3(hist91.crop, raster_name = 'west_1991-2020', main = '1991-2020', save = save)
map_best_geno_crop_k3(fut8.crop, main = '2041-2070 (8 GCM ensemble)', save = save, raster_name = 'west_2041-2070_8GCMs')
map_best_geno_crop_k3(fut13.crop, main = '2041-2070 (13 GCM ensemble)', save = save, raster_name = 'west_2041-2070_13GCMs')

```
