---
title: "Predict growth changes under future climate"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    number_sections: true
    highlight: rstudio
---

Using the genotype-specific modeled reaction norms produced by 'transfer_function_multiyear_linear_mixed_effects_model.Rmd' script, predict how each genotype's growth and survival could change under future temperatures (represented by MCMT).

# Setup

## Load libraries and data

```{r setup}

library(ggplot2)
library(glmmTMB) # model
library(sf) # mapping
library(terra) # climate rasters
library(RColorBrewer) # colors
library(circlize) # colorRamp2

# load garden data - named 'dat'
# remove the 2024 data
load('data/clean/mini_garden_phenotypic_and_climate_data_2021-2024.Rdata')
dat <- dat[, !endsWith(colnames(dat), '2024')]

# load model info
# mod: model
# df.scaled: scaled mini garden data that went into the model
# scaling_factor: used to scale variables before fitting model
load('results/model_prediction/glmTMB_multiyear_model_outputs_garden_MCMT_2021-2022_vs_GrowthIncrement_2021-2022.Rda')

# load predicted heights for each genotype - named 'predictions'
load('results/model_prediction/predictedGrowth_acrossClimates_byGenotype_garden_MCMT_2021-2022.Rdata')
preds <- predictions
rm(predictions)

sessionInfo()

```


```{r markdown_settings}

knitr::opts_chunk$set(fig.width = 8, fig.height = 10)

```

## Back-transform input data

Data used for input to model was scaled - back-transform it to the actual values.

```{r backtransform}

df <- df.scaled

for(n in 1:length(scaling_factor)){
  
  column <- names(scaling_factor)[n]
  
  df[,column] <- df.scaled[,column]/scaling_factor[[n]]
  
}

head(df)

```

## Extract genotype info

Get genetic and climate data for each genotype

```{r genotype_info}

# get info for each genotype to use for extracting climate data and predicting height under future climates

# columns to get - these will be the same for each replicate of a genotype
dput(colnames(dat))
columns <- c("Genotype", "Plant_ID", "Pb", "Pt", "transect", "provenance_latitude", "provenance_longitude", "provenance_elevation_m", "provenance_TD", "provenance_MCMT", "genetic_PC1", "genetic_PC2", "genetic_PC3", "genetic_PC4", "genetic_PC5", "plastid_ID", "color_Pt", "color_k3")

# get genotypes
genos <- unique(dat$Genotype)
genos <- genos[! is.na(genos)]

# get the rows that have the first match of each genotype
rows <- match(genos, dat$Genotype)

# get subset of dataframe with these rows and the info we want for each genotype
geno.info <- dat[rows, columns]
rownames(geno.info) <- geno.info$Genotype

# remove the ones with missing genetic info
geno.info <- geno.info[! is.na(geno.info$genetic_PC1),]
dim(geno.info)
str(geno.info)

```

# Get future climates for each genotype

## Load climate data

Load future and historic MCMT raster files

ClimateNA raster files used here are available from DataBasin:

https://adaptwest.databasin.org/pages/adaptwest-climatena/

Species range shapefiles are from Little 1971 and are available from DataBasin:

P. balsamifera: https://databasin.org/datasets/91380e091ca048359a66fc65962ed210/

P. trichocarpa: https://databasin.org/datasets/84e47784fe2a463c8b292007fb43f2cd/

```{r load_climate}

# which climate variable is being used? 
# can change this to use a different variable
clim_var <- 'MCMT'

# get historic and future climate

hist <- rast(paste('data/climate/climateNA/Normal_1961_1990/Normal_1961_1990_bioclim/Normal_1961_1990_', clim_var, '.tif', sep = ''))
crs(hist, proj = T)

hist91 <- rast(paste('data/climate/climateNA/Normal_1991_2020/Normal_1991_2020_', clim_var, '.tif', sep = ''))
crs(hist, proj = T)

# climateNA provides future ensemble models built from 8 GCMs and 13 GCMs - test both

fut8 <- rast(paste('data/climate/climateNA/future/ensemble_8GCMs_ssp245_2041_2070_bioclim/ensemble_8GCMs_ssp245_2041_2070_', clim_var, '.tif', sep = ''))

fut13 <- rast(paste('data/climate/climateNA/future/ensemble_13GCMs_ssp245_2041_2070_bioclim/ensemble_13GCMs_ssp245_2041_2070_', clim_var, '.tif', sep = ''))

################ convert coords

# convert coordinates of collection sites to the CRS used by climateNA
crs.cna <- crs(hist, proj = T)

# provenance coordinates
coords.prov <- st_as_sf(geno.info[,c("provenance_longitude", "provenance_latitude")], coords = c(1,2), crs = st_crs(4326))
coords.prov.cna <- st_transform(coords.prov$geometry, crs = crs.cna)

# load shapefile of state/province borders
borders <- read_sf('data/shapefiles/NorthAmerica_PoliticalBoundaries_Shapefile/NA_PoliticalDivisions/data/bound_p/boundaries_p_2021_v3.shp')
borders <- st_transform(borders, crs = crs.cna)
# simplify to 1 km
borders <- st_simplify(borders, dTolerance = 1000)

# species rangemaps
# shapefile
balsam <- st_read('data/shapefiles/Pbal_shapefile/data/commondata/data0/popubals.shp')
tricho <- st_read('data/shapefiles/Ptri_shapefile/data/commondata/data0/poputric.shp')

# convert CRS
balsam <- st_transform(balsam, crs.cna)
tricho <- st_transform(tricho, crs.cna)

# test plot
plot(hist)

plot(balsam$geometry, add = T, border = 'navy')
plot(tricho$geometry, add = T, border = 'grey50')
plot(borders$geometry, add = T)



```

## Extract future climates for each genotype's home site

```{r extract_future}

# extract from raster and add to geno.info

# 8 GCM ensemble
tmp <- extract(fut8, y = st_as_sf(coords.prov))
geno.info$future8_MCMT <- tmp[,2]

# 13 GCM ensemble
tmp <- extract(fut13, y = st_as_sf(coords.prov))
geno.info$future13_MCMT <- tmp[,2]

# predictions from the two ensembles are very similar
plot(geno.info$future8_MCMT, geno.info$future13_MCMT)
abline(0, 1, col = 'blue')

```

# Predict height at future climates

## Predict functions

```{r predict_functions}

# originally from transfer function script

# predict each genotype's response to a vector of climate (MCMT) values
# garden_clims is a vector

predict_genotype <- function(model, 
                             type = 'response', # can also be conditional or zprob
                             effects,
                             home_clim,
                             garden_clims){
  
  # vector to save predictions
  pred_height <- vector()
  
  # loop through garden climates
  for(g in 1:length(garden_clims)){
    
    home <- home_clim
    garden <- garden_clims[g]
    
    # make newdata to give to predict()
    new_data <- as.data.frame(cbind(home, garden, home^2, garden^2))
    colnames(new_data) <- c('home_clim', 'garden_clim', 'home_clim_2', 'garden_clim_2')
    # add other random or fixed effects
      for(n in 1:length(effects)){
        new_data[names(effects[n])] <- effects[n]
      }
    
    # first have to scale new_data to match the scaled data used in model
    for(v in 1:length(new_data)){
      var <- names(new_data)[v]
      
      # only scale if var exists in scaling_factor - random effects like genotype are not in scaling factor
      # IF VARIABLES ARE NAMED DIFFERENTLY HERE IN effects AND scaling_factor IT WILL SKIP THEM
      if(var %in% names(scaling_factor)){
        new_data[var] <- new_data[var]*scaling_factor[[var]]
      } else {
        warning(var, ' does not have a scaling factor, skipping')
      }
      
      
    }
    
    pred <- predict(mod, type = type, newdata = new_data, re.form = NA, allow.new.levels=TRUE)
    # back transform from log if we're predicting height
    # zprob is probability of a zero, so don't transform
    if(type %in% c('response', 'conditional')){
      pred <- exp(pred)-1
    }
    pred_height[g] <- pred
    
  }
  
  # save
  pred_height_df <- data.frame(garden_clim = garden_clims, predicted_height = pred_height) 
  
return(pred_height_df)
}


```

## Predict response to historic and future climate for each genotype

```{r predict_hist, warning = F}

# model predictions for historic data
# we already have historic provenance data for the model, so don't need to extract it from the raster

# what climate variable are we using?
clim_var <- 'MCMT'
prov_clim_colname <- paste('provenance', clim_var, sep = '_')
gard_clim_colname <- paste('garden', clim_var, '1961_1990', sep = '_')
fut8_colname <- paste('future8', clim_var, sep = '_')
fut13_colname <- paste('future13', clim_var, sep = '_')


# add predictions to geno.info dataframe
# historic
geno.info$hist_response <- NA
geno.info$hist_height <- NA
geno.info$hist_mortality <- NA

# future 8
geno.info$fut8_response <- NA
geno.info$fut8_height <- NA
geno.info$fut8_mortality <- NA

# future 13
geno.info$fut13_response <- NA
geno.info$fut13_height <- NA
geno.info$fut13_mortality <- NA


for(n in 1:nrow(geno.info)){
  
  geno <- rownames(geno.info)[n]
  
  # setup effects
  g.pc1 <- geno.info$genetic_PC1[n]
  g.pc2 <- geno.info$genetic_PC2[n]
  g.pc3 <- geno.info$genetic_PC3[n]
  g.home_clim <- geno.info[,prov_clim_colname][n]
  
  # garden clims are the historic and future climates
  # in order: home climate (historic), future climate 8 GCMs, future climate 13 GCMs
  g.clims <- c(g.home_clim, geno.info[,fut8_colname][n], geno.info[,fut13_colname][n])
  
  # predict overall response
  resp <- predict_genotype(mod,
                   type = 'response',
                   effects = list(pc1 = g.pc1, 
                                  pc2 = g.pc2,
                                  pc3 = g.pc3,
                                  genotype = geno), 
                   home_clim = g.home_clim, 
                   garden_clims = g.clims)
  
  # save to df
  geno.info$hist_response[n] <- resp$predicted_height[1]
  geno.info$fut8_response[n] <- resp$predicted_height[2]
  geno.info$fut13_response[n] <- resp$predicted_height[3]

  
  
  # predict height (conditional model)
  cond <- predict_genotype(mod,
                           type = 'conditional',
                           effects = list(pc1 = g.pc1, 
                                          pc2 = g.pc2,
                                          pc3 = g.pc3,
                                          genotype = geno), 
                           home_clim = g.home_clim, 
                           garden_clims = g.clims)
  
  # save to df
  geno.info$hist_height[n] <- cond$predicted_height[1]
  geno.info$fut8_height[n] <- cond$predicted_height[2]
  geno.info$fut13_height[n] <- cond$predicted_height[3]

  # predict probability of mortality (zero-inflated model)
  zi <- predict_genotype(mod,
                         type = 'zprob',
                         effects = list(pc1 = g.pc1, 
                                        pc2 = g.pc2,
                                        pc3 = g.pc3,
                                        genotype = geno), 
                         home_clim = g.home_clim, 
                         garden_clims = g.clims)
  # save to df
  geno.info$hist_mortality[n] <- zi$predicted_height[1]
  geno.info$fut8_mortality[n] <- zi$predicted_height[2]
  geno.info$fut13_mortality[n] <- zi$predicted_height[3]

  
 #cat(paste('done with', n, '\n'))
  
}

# predict_genotype() gives warning:
# '## Warning in predict_genotype(mod, type = "response", effects = list(pc1 = g.pc1,
## : genotype does not have a scaling factor, skipping'

# this is fine, genotype is a string not numeric and shouldn't be scaled

```

## Calculate change in fitness under future climates

```{r calc_change}

# calculate changes in fitness under future climates

##########
# future (8 GCMs)

# net change in response
geno.info$fut8_response_change <- geno.info$fut8_response - geno.info$hist_response
# percent change
geno.info$fut8_response_change_percent <- (geno.info$fut8_response - geno.info$hist_response)/geno.info$hist_response

# net change in height
geno.info$fut8_height_change <- geno.info$fut8_height - geno.info$hist_height
# percent change
geno.info$fut8_height_change_percent <- (geno.info$fut8_height - geno.info$hist_height)/geno.info$hist_height

# net change in mortality probability
geno.info$fut8_mortality_change <- geno.info$fut8_mortality - geno.info$hist_mortality
# percent change
geno.info$fut8_mortality_change_percent <- (geno.info$fut8_mortality - geno.info$hist_mortality)/geno.info$hist_mortality

##########
# future (13 GCMs)

# net change in response
geno.info$fut13_response_change <- geno.info$fut13_response - geno.info$hist_response
# percent change
geno.info$fut13_response_change_percent <- (geno.info$fut13_response - geno.info$hist_response)/geno.info$hist_response

# net change in height
geno.info$fut13_height_change <- geno.info$fut13_height - geno.info$hist_height
# percent change
geno.info$fut13_height_change_percent <- (geno.info$fut13_height - geno.info$hist_height)/geno.info$hist_height

# net change in mortality probability
geno.info$fut13_mortality_change <- geno.info$fut13_mortality - geno.info$hist_mortality
# percent change
geno.info$fut13_mortality_change_percent <- (geno.info$fut13_mortality - geno.info$hist_mortality)/geno.info$hist_mortality

# preliminary plots
par(mfrow = c(2,3))
hist(geno.info$fut8_response_change_percent)
hist(geno.info$fut8_height_change_percent)
hist(geno.info$fut8_mortality_change_percent)

hist(geno.info$fut13_response_change_percent)
hist(geno.info$fut13_height_change_percent)
hist(geno.info$fut13_mortality_change_percent)

```

# Map predictions for each collected genotype/site

```{r map_local_response_one_var}

# baseline plot of climate difference

diff13 <- fut13 - hist
diff8 <- fut8 - hist

par(mfrow = c(1,2))
plot(diff13)
plot(diff8)

# color palette for MCMT base layer

colf <- colorRampPalette(brewer.pal(9, 'YlOrRd'))
cols <- colf(100)
cols2 <- adjustcolor(cols, alpha = 0.5)

# function to map the response 

# future = the raster object for change in climate to use as background map
# response = the column name from geno.info of the response to future climate to plot
# main = plot title
# rev = reverse response color palette? Can use for mortality so a negative/positive response has the same color
map_change <- function(future, response, main = '', rev = FALSE, legend = FALSE, raster = FALSE, ...){
  
  # which model response are we plotting?
  resp_name <- response
  
  colf <- colorRampPalette(brewer.pal(9, 'YlOrRd'), alpha = 0.5)
  cols <- colf(100)
  cols <- adjustcolor(cols, alpha = 0.6)
  
  #par(mfrow = c(1,1))
  plot(future, 
       col = ifelse(raster == TRUE, cols, 'white'),
       main = main,
       xaxt = 'n', yaxt = 'n', 
       pax = list(side=NA),
       xlim = c(-3000000,-500000), ylim = c(400000,3200000), 
       legend = legend,
       plg = list(title = 'Future -\nHistoric\n MCMT (°C)'),
       ...)
  
  # add borders and species ranges
  plot(borders$geometry, add = T, lwd = 0.5)
  plot(balsam$geometry, add = T, border = 'navy')
  plot(tricho$geometry, add = T, border = 'darkgreen')
  

  # color palette for change in response
  min_change <- min(geno.info[,resp_name])
  max_change <- max(geno.info[,resp_name])
  
  ceil2 <- max(abs(min_change), abs(max_change))
  
  if(rev == FALSE){
    colf <- colorRamp2( breaks = c(-ceil2, 0, ceil2), colors = c('darkorchid4', 'white', 'palegreen4'))
  } else if(rev == TRUE){
    colf <- colorRamp2( breaks = c(-ceil2, 0, ceil2), colors = c('palegreen4', 'white', 'darkorchid4'))
  }
  
  
  cols2 <- colf(geno.info[,resp_name])
  
  # plot points for each genotype
  points(coords.prov.cna, bg = cols2,  pch = 21, cex = 2, lwd = 1.5)
  
  # add colors of upper and lower bounds to legend
  if(rev == FALSE){
     pt.cols <- c('darkorchid4', 'white', 'palegreen4')
  } else if(rev == TRUE) {
    pt.cols = c('palegreen4', 'white', 'darkorchid4')
  }
 
  
  # legend(-2.8e6,1.5e6, title = '% Change',
  #        legend = c(round(-ceil2,2), 0, round(ceil2,2)),
  #        pch = 21,
  #        pt.bg = pt.cols,
  #        pt.cex = 2)
  
  vals <- round(c(min_change, 0, max_change),2)
  # if range doesn't overlap 0, remove it
  if(min_change * max_change > 0){
    vals <- vals[vals!=0]
  }
  
  legend(-2.8e6,1.5e6, title = '% Change',
         legend = vals,
         pch = 21,
         pt.bg = colf(vals),
         pt.cex = 2)
  
  
}


par(mfrow = c(2,3))
map_change(diff13, 'fut13_response_change_percent')
map_change(diff13, 'fut13_height_change_percent')
map_change(diff13, 'fut13_mortality_change_percent', rev = T)

map_change(diff8, 'fut8_response_change_percent')
map_change(diff8, 'fut8_height_change_percent')
map_change(diff8, 'fut8_mortality_change_percent', rev = T)

# raw values (cm or % mortality)
# legend title not correct here
par(mfrow = c(2,3))
map_change(diff13, 'fut13_response_change')
map_change(diff13, 'fut13_height_change')
map_change(diff13, 'fut13_mortality_change', rev = T)

map_change(diff8, 'fut8_response_change')
map_change(diff8, 'fut8_height_change')
map_change(diff8, 'fut8_mortality_change', rev = T)


#png(file = paste('results/model_prediction/fitness_changes_map_2yearModel', clim_var, '_13GCMs.png', sep = ''), height = 8, width = 8, res = 300, units = 'in')


# lighter colors for temperature raster
colf <- colorRampPalette(brewer.pal(9, 'YlOrRd'))
cols <- colf(100)
cols2 <- adjustcolor(cols, alpha = 0.5)

# plot genotypes and the 13-gcm ensemble
par(mfrow = c(2,2))
# set margins
marg <- c(1,1,4,3)

plot(diff13, col = cols2, 
     main = paste(clim_var, ' 2041-2070, 13 GCMs', sep = ''),
     xaxt = 'n', yaxt = 'n', 
     pax = list(side=NA),
     mar = marg,
     xlim = c(-3000000,-500000), ylim = c(400000,3200000), 
     plg = list(title = 'Future -\nHistoric\n MCMT (°C)'))
# add borders and species ranges
plot(borders$geometry, add = T, lwd = 0.5)
plot(balsam$geometry, add = T, border = 'navy')
plot(tricho$geometry, add = T, border = 'darkgreen')

points(coords.prov.cna, bg = geno.info$color_Pt, pch = 21, cex = 2)

map_change(diff13, 'fut13_response_change_percent', main = 'Overall response', mar = marg)
map_change(diff13, 'fut13_height_change_percent', main = 'Growth', mar = marg)
map_change(diff13, 'fut13_mortality_change_percent', rev = T, main = 'Probability of Mortality', mar = marg)


#dev.off()


```

# Map rangewide predictions for each genotype

## Map past and future fitness predictions and change for each genotype

```{r map_rangewide_genotype_one_var, fig.height = 5, fig.width = 12}


# custom predict function - predict height from discrete set of climate values that have already been calculated

predict_disc <- function(value, pred_clim, pred_value){
  if(! is.na(value)){
    # which value in pred_clim (already predicted climates) is closest to the value we want to predict?
    # returns the index
    ind <- which.min(abs(pred_clim - value))
    
    # return the predicted height for that index
    return(pred_value[ind])
    
  } else if(is.na(value)){
    return(NA)
  }
}

# which climate variable to use?
clim_var <- 'MCMT'
hist_rast <- hist
fut_rast <- fut13

# set up input raster - historic climate
hist.crop <- crop(hist_rast, ext(-4e6, 0, 0, 3e6))
plot(hist.crop)
# make values outside predict range NA
# need pred_clim - this is the same for all genotypes, so can just use first one
#pred_clim <- preds$overall$genotype_210$garden_clim
#pred_clim <- dat[,gard_clim_colname]
pred_clim <- c(dat$garden_MCMT_2021, dat$garden_MCMT_2022)

msk <- ifel(hist.crop > max(pred_clim) | hist.crop < min(pred_clim), NA, 1)
hist.crop <- mask(hist.crop, msk)
plot(hist.crop, colNA = 'grey')

# future climate
fut13.crop <- crop(fut_rast, ext(-4e6, 0, 0, 3e6))
plot(fut13.crop)
# make values outside predict range NA
# need pred_clim - this is the same for all genotypes, so can just use first one
# pred_clim <- preds$overall$genotype_210$garden_clim
msk <- ifel(fut13.crop > max(pred_clim) | fut13.crop < min(pred_clim), NA, 1)
fut13.crop <- mask(fut13.crop, msk)
plot(fut13.crop, colNA = 'grey')


# loop through genotypes, predict height across range

par(mfrow = c(1,3), mar = c(5,4,4,5))
for(n in 1:length(preds$overall)){
  
  # predicted climate and predicted height for this genotype
  pred_clim <- preds$overall[[n]]$garden_clim
  pred_value <- preds$overall[[n]]$predicted_height
  
  #### historic
  # set heights for each raster cell in historic climate (returns a list)
  heights_cells_hist <- lapply(values(hist.crop), FUN = function(x) predict_disc(x, pred_clim, pred_value))
  # assign heights to raster
  height_hist <- setValues(hist.crop, values = unlist(heights_cells_hist))
  
  # plot
   colf <- colorRampPalette(rev(brewer.pal(9, 'YlGn')))
   cols <- rev(colf(100))
  
  plot(height_hist, col = cols, main = paste(names(preds$overall)[n], '\nFitness in Historic Climate', sep = ''), colNA = 'grey90')
  
  plot(borders$geometry, add = T)
  plot(balsam$geometry, add = T, border = 'navy')
  plot(tricho$geometry, add = T, border = 'grey50')
  
  points(coords.prov.cna[n], col = 'mediumpurple', lwd = 2, cex = 1.5)
  #points(coords.prov.cna)
  
    #### future
  # set heights for each raster cell in historic climate (returns a list)
  heights_cells_fut <- lapply(values(fut13.crop), FUN = function(x) predict_disc(x, pred_clim, pred_value))
  # assign heights to raster
  height_fut <- setValues(fut13.crop, values = unlist(heights_cells_fut))
  
  # plot

  plot(height_fut, col = cols, main = 'Fitness in Future Climate', colNA = 'grey90')
  
  plot(borders$geometry, add = T)
  plot(balsam$geometry, add = T, border = 'navy')
  plot(tricho$geometry, add = T, border = 'grey50')
  
  points(coords.prov.cna[n], col = 'mediumpurple', lwd = 2, cex = 1.5)
  #points(coords.prov.cna)
  
  ### plot difference
  height_diff <- height_fut - height_hist
  
  # min and max values for diverging color palette
  min_var <- minmax(height_diff)[1,]
  max_var <- minmax(height_diff)[2,]
  ceil <- max(abs(min_var), abs(max_var))
  
  colf_diff <- colorRampPalette(brewer.pal(11, 'RdBu'))
  cols_diff <- colf_diff(100)
  
  plot(height_diff, col = cols_diff, range = c(-ceil, ceil), main = 'Fitness change', colNA = 'grey90')
  
  plot(borders$geometry, add = T)
  plot(balsam$geometry, add = T, border = 'navy')
  plot(tricho$geometry, add = T, border = 'grey50')
  
  points(coords.prov.cna[n], col = 'black', lwd = 2, cex = 1.5)
  
  print(paste('done with', n))
  
}

```

