---
title: "Predict height in maxis"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    number_sections: true
    highlight: rstudio
---

Test whether model built from mini gardens dataset can accurately predict growth in the maxi gardens (3 sites, one of which is novel, and 544 genotypes, 500 of which are novel)

This code is mostly pulled from 'code/transfer_function_multiyear_linear_mixed_effects_model.Rmd' script

# Setup

## packages and data

```{r setup}

library(ggplot2)
library(sjPlot) # nice plots of models
library(glmmTMB) #zero-inflated  model and also used with sjPlot
library(bbmle) # for AICtab

# maxi garden data - named 'maxis'
load('data/clean/maxi_garden_phenotypic_climate_genetic_data_2021-2023.Rdata')
str(maxis)

# load model info
# mod: model
# df.scaled: scaled mini garden data that went into the model
# scaling_factor: used to scale variables before fitting model
load('results/model_prediction/glmTMB_multiyear_model_outputs_garden_MCMT_2021-2022_vs_GrowthIncrement_2021-2022.Rda')

```

```{r markdown_settings}

knitr::opts_chunk$set(fig.width = 10, fig.height = 10)

```

```{r compare_datasets}

# which genotypes are in maxis but not minis?
genoMinis <- unique(df.scaled$genotype)
genoMaxis <- unique(maxis$TAG)


length(genoMinis)
length(genoMaxis)

sum(! genoMaxis %in% genoMinis) # 500

```

# clean up data

## change ND name
```{r cleanup_ND}

# rename ND to NDSU to match mini garden dataset
maxis$garden <- gsub('ND', 'NDSU', maxis$garden)


```

Cleanup using same methods as in model fitting script 'code/transfer_function_linear_mixed_effects_model.Rmd'


## set dead to 0

```{r cleanup_set_dead_to_0}

# dead trees should have a growth of zero if they were dead the whole year
# some trees died during the season - keep these growth values since it's real growth and should be related to fitness

# change these growth values from NA to 0
par(mfrow = c(1,2))
hist(maxis$GrowthIncrement_2021)
maxis[which(maxis$LIVE_Y_N_2021 == 0 & is.na(maxis$GrowthIncrement_2021)), 'GrowthIncrement_2021'] <- 0
hist(maxis$GrowthIncrement_2021)

hist(maxis$GrowthIncrement_2022)
maxis[which(maxis$LIVE_Y_N_2022 == 0 & is.na(maxis$GrowthIncrement_2022)), 'GrowthIncrement_2022'] <- 0
hist(maxis$GrowthIncrement_2022)

# save a list of which trees are alive each year
# used later to exclude dead trees when evaluating conditional model
alive21 <- rownames(maxis)[maxis$LIVE_Y_N_2021 == 1]
alive22 <- rownames(maxis)[maxis$LIVE_Y_N_2022 == 1]


```

```{r cleanup_remove_dead, eval = F, include = F}

# remove trees that were dead at the end of 2021
# need to modify this if using across multiple years
dim(maxis)
maxis <- maxis[which(maxis$LIVE_Y_N_2021 == 1),]
dim(maxis)

```

## remove negative heights

```{r cleanup_remove_negative_heights}

# remove negative heights - these are either measurement error or being eaten

# 2021
hist(maxis$GrowthIncrement_2021)
sum(maxis$GrowthIncrement_2021 < 0, na.rm = T) # 65

maxis[which(maxis$GrowthIncrement_2021 <0), 'GrowthIncrement_2021'] <- NA

# 2022
hist(maxis$GrowthIncrement_2022)
sum(maxis$GrowthIncrement_2022 < 0, na.rm = T) # 47

maxis[which(maxis$GrowthIncrement_2022 <0), 'GrowthIncrement_2022'] <- NA

# 2023
hist(maxis$GrowthIncrement_2023)
sum(maxis$GrowthIncrement_2023 < 0, na.rm = T) # 50

maxis[which(maxis$GrowthIncrement_2023 <0), 'GrowthIncrement_2023'] <- NA


# check
hist(maxis$GrowthIncrement_2021)
hist(maxis$GrowthIncrement_2022)
hist(maxis$GrowthIncrement_2023)

```

# Set up model

## choose model variables

Use MCMT as climate variable

```{r setup_choose_model_variables}

# MCMT vs height

# climate
garden_clim_colname <- 'garden_MCMT_2021'
home_clim_colname <- 'provenance_MCMT_1961_1990'
clim_label <- 'MCMT'

# phenotype
pheno_colname <- 'GrowthIncrement_2021'
pheno_label <- 'Growth Increment 2021'

###############################################

# setup variables and put into a dataframe

# set climate and phenotype variables
garden_clim <- maxis[,garden_clim_colname]
garden_clim_2 <- maxis[,garden_clim_colname]^2
home_clim <- maxis[,home_clim_colname]
home_clim_2 <- maxis[,home_clim_colname]^2
pheno <- maxis[,pheno_colname]

# random effects

# need to edit block so the VA and ND gardens don't have same block names?
# add 'maxi' to block name to distinguish from the NDSU and VA mini gardens blocks
# otherwise the same random effect will be applied even though they are actually different blocks
block <- paste('maxi', as.character(interaction(maxis$garden, maxis$BLOCK, drop = T)), sep = '.')

Pt <- maxis$k2_tricho
genotype <- as.character(maxis$TAG)
garden <- as.character(maxis$garden)
pc1 <- maxis$genetic_PC1
pc2 <- maxis$genetic_PC2
pc3 <- maxis$genetic_PC3
pc4 <- maxis$genetic_PC4
pc5 <- maxis$genetic_PC5

# put in df
df <- data.frame(pheno, garden_clim, garden_clim_2, home_clim, home_clim_2, Pt, genotype, garden, block, pc1, pc2, pc3, pc4, pc5)
rownames(df) <- rownames(maxis)

# add year and individual
df$year <- 2021
df$indiv <- rownames(maxis)

str(df)


```

## scale variables

```{r setup_scale}

# subset to the variables that were used in original model
# get variables that were put into model
# variables have to be named the same in this script as in 'choose_model_variables' chunk
vars <- colnames(mod$frame)
# rename response variable
vars[1] <- 'pheno'

df <- df[,vars]
str(df)


# scale using scaling factors
# go through each column and check if it exists in scaling_factor
# skips non-numeric columns

df.scaled <- df

for(n in 1:ncol(df)){
  
  if(colnames(df)[n] %in% names(scaling_factor)){
    # to scale, multiply by scaling factor
    df.scaled[,n] <- df.scaled[,n] * as.numeric(scaling_factor[colnames(df)[n]])
  }
}

head(df.scaled)  


```

# Predict growth and mortality from model

## Functions

```{r functions}

# modified from 'code/transfer_function_linear_mixed_effects_model.Rmd'

# function to calculate correlation between predicted and observed values for a model (essentially the R value)

# mod is the model
# it pulls the dataset from the model object, which allows us to remove rows with NAs for just the variables used in this model
# re.form argument goes to predict(); to include random effects set to NULL, to set random effects to zero set to NA

model_R <- function(mod, newdata, type = 'response', re.form = NULL, se.fit = FALSE){
  
  # setup
  # include random effects in predictions or set random effects to zero?
  # this is the re.form argument in predict()
  #re.form <- ifelse(include_REs == TRUE, eval(NULL), NA)
  
  df <- newdata
  
  # make new dataframe with no NA values
  noNA <- df[complete.cases(df),]

  
  # predict phenotype (generally height)
  noNA$pred <- predict(mod, newdata = noNA, type = type, re.form = re.form, se.fit = se.fit)
  
  # un-log-transform
  noNA$pred <- exp(noNA$pred) - 1
  
  # if type = conditional, remove dead trees (which are modeled separately in zero-inflated model)
  if(type == 'conditional'){
    noNA <- noNA[alive21,]
  }
  
  # calculate correlation
  cor <- cor.test(noNA$pheno, noNA$pred)
  
  return(list(actual = noNA$pheno, predicted = noNA$pred, cor = cor, data_noNAs = noNA))
  
}

# test
# cor <- model_R(mod, re.form = NULL)

# plot which takes output of model_R

plot_predicted_vs_actual <- function(input, title = NULL, col = rgb(0,0,0,0.1), col_1to1 = 'red', col_fit = 'blue', legend = TRUE){
  
  pval <- round(input$cor$p.value, 4)
  cor_est <- as.numeric(round(input$cor$estimate, 3))
  main <- paste(title, '\n', 'R = ', cor_est, ' | p = ', pval, sep = '')
  
  plot(input$actual, input$predicted, 
       pch = 16, 
       col = col, 
       xlab = "Actual growth increment (cm)", 
       ylab = "Predicted growth increment (cm)",
       title(main, adj = 0))
  abline(0, 1, col = col_1to1, lty = 2) # 1to1 line
  abline(lm(input$predicted ~ input$actual), col = col_fit) # fit line
  
  # add text
  if(legend == TRUE){
    legend('bottomright', fill = c(col_1to1, col_fit), legend = c('1:1 line', 'Best fit'))
  }
}



```


## Predict for maxi garden data

```{r predict, warning = FALSE}

# predict growth increment for each tree in maxi gardens, based on genetic PCs, home climate, and garden climate, and compare predictions to actual values

# see ?predict.glmmTMB

# This will give a warning about new random effects, because some genotypes and gardens were not included in training dataset (mini gardens)

# Warning messages:
# 1: In checkTerms(data.tmb1$terms, data.tmb0$terms) :
#   Predicting new random effect levels for terms: 1 | genotype, 1 | block:garden, 1 | garden, 1 | indiv
# Disable this warning with 'allow.new.levels=TRUE'

# 2: In checkTerms(data.tmb1$termszi, data.tmb0$termszi) :
#   Predicting new random effect levels for terms: 1 | genotype, 1 | block:garden, 1 | garden, 1 | indiv
# Disable this warning with 'allow.new.levels=TRUE

par(mfrow = c(2,2))
# overall model
plot_predicted_vs_actual(model_R(mod, newdata =  df.scaled, re.form = NULL), title = 'overall, with random effects')
plot_predicted_vs_actual(model_R(mod, newdata = df.scaled, re.form = NA), title = 'overall, no random effects')
# conditional model
plot_predicted_vs_actual(model_R(mod, type = 'conditional', newdata =  df.scaled, re.form = NULL), title = 'conditional, with random effects')
plot_predicted_vs_actual(model_R(mod, type = 'conditional', newdata = df.scaled, re.form = NA), title = 'conditional, no random effects')


# for each garden

# overall response
#png(file = paste('results/model_prediction/maxi_growth_predictions_vs_actual_from_model_2yearModel_', garden_clim_colname, '.png', sep = ''),
# height = 8,
# width = 12,
# res = 300,
# units = 'in')

par(mfrow = c(3,4))
gards <- unique(df.scaled$garden)

for(n in 1:length(gards)){
  
  gard <- gards[n]
  # overall model
  plot_predicted_vs_actual(model_R(mod, df.scaled[df.scaled$garden == gard,], re.form = NULL),
                           title = paste(gard, 'overall, with random effects'),
                           legend = ifelse(n == 1, T, F))
  plot_predicted_vs_actual(model_R(mod, df.scaled[df.scaled$garden == gard,], re.form = NA), 
                           title = paste(gard, 'overall, no random effects'), legend = F)
  # conditional model
    plot_predicted_vs_actual(model_R(mod, df.scaled[df.scaled$garden == gard,], re.form = NULL, type = 'conditional'),
                           title = paste(gard, 'conditional, with random effects'), legend = F)
  plot_predicted_vs_actual(model_R(mod, df.scaled[df.scaled$garden == gard,], re.form = NA, type = 'conditional'), 
                           title = paste(gard, 'conditional, no random effects'), legend = F)
  
}

#dev.off()


# just conditional with random effects, to match the mini gardens validation plots

#png(file = paste('results/model_prediction/validation_MaxiPredictions_2yearModel', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''), width = 9, height = 3, res = 300, units = 'in')

cols <- c("#2B3D26", "#B3446C", "#2b3e85")

par(mfrow = c(1,3))

for(n in 1:length(gards)){
  
  gard <- gards[n]
  col <- adjustcolor(cols[n], alpha = 0.6)
  
  plot_predicted_vs_actual(model_R(mod, df.scaled[df.scaled$garden == gard,], re.form = NA, type = 'conditional'), 
                           title = paste(gard, '(Maxi)'),
                           col = col,
                           col_fit = col,
                           col_1to1 = 'grey')
  
}

#dev.off()
  
    
    

# test prediction ability for each genotype

genos <- unique(df.scaled$genotype)


geno.r.resp <- rep(NA, length(genos))
geno.r.cond <- rep(NA, length(genos))


for(n in 1:length(genos)){
  
  geno <- genos[n]
  
  sub <- df.scaled[df.scaled$genotype == geno,]
  
  # checks before trying to predict
  # 1. need to have at least three observations
  # 2. if genotype is missing information, skip it
  # ok if phenotype has some NAs
  if(length(sub$pheno[! is.na(sub$pheno)]) > 2 &
     sum(is.na(sub[,colnames(sub != 'pheno')])) == 0){
    
    resp <- model_R(mod, sub, type = 'response', re.form = NULL)
    
    # only run conditional if there are enough live trees
    if(nrow(sub[rownames(sub) %in% alive21,]) > 2){
       cond <- model_R(mod, sub, type = 'conditional', re.form = NULL)
    }
   
    
    geno.r.resp[n] <- resp$cor$estimate
    geno.r.cond[n] <- cond$cor$estimate
    
  }
  
  
  # if(n %% 20 == 0){
  #   cat(paste('done with', n, '\n'))
  # }
}

names(geno.r.resp) <- genos
names(geno.r.cond) <- genos

par(mfrow = c(1,2))
hist(geno.r.resp)
hist(geno.r.cond)


# which genotypes are more predictable?

geno_info <- as.data.frame(matrix(nrow= length(genos), ncol = 6))
colnames(geno_info) <- c('k2_tricho', 'genetic_PC1', 'genetic_PC2', 'genetic_PC3', 'TRANSECT', 'in_minis' )

rownames(geno_info) <- genos

geno_colnames <- c('k2_tricho', 'genetic_PC1', 'genetic_PC2', 'genetic_PC3', 'TRANSECT')

for(n in 1:nrow(geno_info)){
  
  geno <- rownames(geno_info)[n]
  
  # get first row with info for this genotype
  geno_row <- maxis[match(geno, maxis$TAG), geno_colnames]
  
  geno_info[n, geno_colnames] <- geno_row
  
  # is the genotype in the mini gardens?
  geno_info[n, 'in_minis'] <- geno %in% mod$frame$genotype

}

geno_info$Rval_resp <- geno.r.resp
geno_info$Rval_cond <- geno.r.cond

# there are a lot of missing values due to low sample size for some genotypes. Look at using residuals instead? or some other measure of accuracy?

# are predictions better for genotypes that were included in the minis (training dataset)?
# average R values are similar, but more variance in accuracy for non-mini genotypes, and most of the negative R values are non-mini genotypes
par(mfrow = c(1,2), mar = c(8,4,4,3))
boxplot(geno_info$Rval_resp ~ geno_info$in_minis)
abline(h = 0, col = 'blue')
boxplot(geno_info$Rval_cond ~ geno_info$in_minis)
abline(h = 0, col = 'blue')

plot(density(geno_info[geno_info$in_minis == F,'Rval_resp'], na.rm = T))

# is there variance across transects?
# not too much variance across transects, including WY which was not in minis
# lost of variation in accuracy for crowsnest conditional model
par(mfrow = c(1,2), mar = c(8,4,4,3))
boxplot(geno_info$Rval_resp ~ geno_info$TRANSECT, las = 2, xlab = '')
abline(h = 0, col = 'blue')
boxplot(geno_info$Rval_cond ~ geno_info$TRANSECT, las = 2, xlab = '')
abline(h = 0, col = 'blue')

# plot cor against genetic info
# genotypes with low R values are both balsam and tricho

par(mfrow = c(1,2))

plot(geno_info$k2_tricho, geno_info$Rval_resp)
abline(h = 0, col = 'blue')
plot(geno_info$k2_tricho, geno_info$Rval_cond)
abline(h = 0, col = 'blue')

plot(geno_info$genetic_PC1, geno_info$Rval_resp)
abline(h = 0, col = 'blue')
plot(geno_info$genetic_PC1, geno_info$Rval_cond)
abline(h = 0, col = 'blue')

plot(geno_info$genetic_PC2, geno_info$Rval_resp)
abline(h = 0, col = 'blue')
plot(geno_info$genetic_PC2, geno_info$Rval_cond)
abline(h = 0, col = 'blue')

plot(geno_info$genetic_PC3, geno_info$Rval_resp)
abline(h = 0, col = 'blue')
plot(geno_info$genetic_PC3, geno_info$Rval_cond)
abline(h = 0, col = 'blue')

```
