---
title: "Predict genotype-specific reaction norms"
author: "Alayna Mead"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: true
    number_sections: true
    highlight: rstudio
---

Test the effects of home climate, garden climate, and genetics on the growth and mortality responses of hybrid poplars across different common gardens, then predict genotype-specific reaction norms to MCMT. Uses glmmTMB to fit a zero-inflated mixed-effects model.

# Setup

## data and packages
```{r setup}

library(ggplot2)
library(colorRamp2)
library(RColorBrewer)
library(lme4)
library(sjPlot) # nice plots of models
library(glmmTMB) #zero-inflated  model and also used with sjPlot
library(bbmle) # for AICtab
library(patchwork) # plot ggplots together
library(ggpubr) # easily add pvals to ggplots
library(performance) # for rmse (root mean squared error function)
library(psych) # for pairs.panels()

# using the merged 2024 dataset because Idaho wasn't included in the old 2023 merged file
# remove the 2024 data
load('data/clean/mini_garden_phenotypic_and_climate_data_2021-2024.Rdata')
dat <- dat[, !endsWith(colnames(dat), '2024')]

str(dat)

# print session info
sessionInfo()

# ggplot settings
theme_set(theme_bw() + theme(text=element_text(size=16)))

```

```{r markdown_settings}

knitr::opts_chunk$set(fig.width = 12, 
                      fig.height = 10)

```

## Clean up input data

Set the growth increment to 0 for all trees that died and didn't have a growth increment value. This is needed to include both growth and mortality measures within the same model; it is zero-inflated to evaluate the zeros arising from mortality separately.

```{r cleanup_set_dead_to_0}

# dead trees should have a growth of zero if they were dead the whole year
# some trees died during the season and their height was still recorded - keep these  values since it represents real growth and should be related to fitness


# how many of these trees are there?
length(which(dat$Survival_09_2021 == 0 & is.na(dat$GrowthIncrement_2021))) # 336
length(which(dat$Survival_09_2022 == 0 & is.na(dat$GrowthIncrement_2022))) # 324
length(which(dat$Survival_09_2023 == 0 & is.na(dat$GrowthIncrement_2023))) # 337

# change these growth values from NA to 0
dat[which(dat$Survival_09_2021 == 0 & is.na(dat$GrowthIncrement_2021)), 'GrowthIncrement_2021'] <- 0
dat[which(dat$Survival_09_2022 == 0 & is.na(dat$GrowthIncrement_2022)), 'GrowthIncrement_2022'] <- 0
dat[which(dat$Survival_09_2023 == 0 & is.na(dat$GrowthIncrement_2023)), 'GrowthIncrement_2023'] <- 0


# save a list of which trees are alive at the end of each year
# used later to exclude dead trees when evaluating conditional model
alive21 <- rownames(dat)[dat$Survival_09_2021 == 1 & ! is.na(dat$Survival_09_2021)]
alive22 <- rownames(dat)[dat$Survival_09_2022 == 1 & ! is.na(dat$Survival_09_2022)]
alive23 <- rownames(dat)[dat$Survival_09_2023 == 1 & ! is.na(dat$Survival_09_2023)]

nrow(dat)
length(alive21)
length(alive22)
length(alive23)

# how many trees have values of zero growth?
sum(dat$GrowthIncrement_2021 == 0, na.rm = T) # 381
sum(dat$GrowthIncrement_2022 == 0, na.rm = T) # 343
sum(dat$GrowthIncrement_2023 == 0, na.rm = T) # 370

# total number of trees that are not NA
sum(! is.na(dat$GrowthIncrement_2021)) # 1476
sum(! is.na(dat$GrowthIncrement_2022)) # 1234
sum(! is.na(dat$GrowthIncrement_2023)) # 1133


# note: 'total_growth_increment_2021_2022' has already been calculated in cleanup script. The NAs for dead trees mean that the total growth increment will also (correctly) be an NA if the tree died during one of the years. Don't recalculate total growth increment here - if growth for one year was a 0 because it was dead, multi-year growth would sum to an actual number and would not be left out of the analysis, as it is when the value is NA.
# total_growth_increment summed across years is no longer used in this script, but I'll leave this note here in case that changes

```

```{r cleanup_remove_NA_genotypes}

# remove NA genotypes

dat <- dat[! is.na(dat$Genotype),]

table(dat$Genotype) 
length(table(dat$Genotype)) # 47 genotypes

```

## Save input data

```{r save_input_data}

# save(dat, file = 'data/clean/mini_garden_phenotypic_and_climate_data_2021-2023_deadHeight0_removeNegativeGrowth.Rdata')
# write.csv(dat, file = 'data/clean/mini_garden_phenotypic_and_climate_data_2021-2023_deadHeight0_removeNegativeGrowth.csv')

```


# Functions

```{r functions}

# function to calculate correlation between predicted and observed values for a model (essentially the R value)

# mod is the model
# it pulls the dataset from the model object, which allows us to remove rows with NAs for just the variables used in this model
# re.form argument goes to predict(); to include random effects set to NULL, to set random effects to zero set to NA
# 'alive' is the list of individual trees that are alive and needs to be defined in the environment - it's set up in 'setup_choose_model_variables' chunk

model_R <- function(mod, type = 'response', re.form = NULL, se.fit = FALSE){
  
  # get dataframe with all variables used in model
  df <- mod$frame
  
  # first column is phenotype
  colnames(df)[1] <- 'pheno'
  
  # make new dataframe with no NA values
  # not necessary when using dataframe from model output
  #noNA <- df[complete.cases(df),]

  
  # predict phenotype (generally height)
  df$pred <- predict(mod, type = type, re.form = re.form, se.fit = se.fit)
  
  # un-log-transform height
  df$pred <- exp(df$pred) - 1
  df$pheno <- exp(df$pheno) - 1
  
  # if type = conditional, remove dead trees (which are modeled separately in zero-inflated model)
  if(type == 'conditional'){
    df <- df[alive,]
  }
  
  # calculate correlation between actual and predicted values
  cor <- cor.test(df$pheno, df$pred)
  
  return(list(actual = df$pheno, predicted = df$pred, cor = cor, data_noNAs = df))
  
}

# plot actual vs predicted values and prints the correlation and p-value using the output of model_R as 'input' argument

plot_predicted_vs_actual <- function(input, title = NULL, col = rgb(0,0,0,0.1), col_1to1 = 'red', col_fit = 'blue', legend = TRUE){
  
  pval <- round(input$cor$p.value, 4)
  cor_est <- as.numeric(round(input$cor$estimate, 3))
  main <- paste(title, '\n', 'R = ', cor_est, ' | p = ', pval, sep = '')
  
  plot(input$actual, input$predicted, 
       pch = 16, 
       col = col, 
       xlab = "Height (actual)", 
       ylab = "Height (predicted)",
       title(main, adj = 0))
  abline(0, 1, col = col_1to1, lty = 2) # 1to1 line
  abline(lm(input$predicted ~ input$actual), col = col_fit) # fit line
  
  # add text
  if(legend == TRUE){
    legend('bottomright', fill = c(col_1to1, col_fit), legend = c('1:1 line', 'Best fit'))
  }
}


#######
# for plotting actual mortality vs mortality probability - need logistic model
# this reports the correct statistics but the plot function just uses a linear line - should be changed to a binomial model
# but this is okay for a quick look


model_R_mortality <- function(mod, re.form = NULL, se.fit = TRUE){
  
  # setup
  # include random effects in predictions or set random effects to zero?
  # this is the re.form argument in predict()
  #re.form <- ifelse(include_REs == TRUE, eval(NULL), NA)
  
  df <- mod$frame
  
  # first column is phenotype
  colnames(df)[1] <- 'pheno'
  
  # make new dataframe with no NA values
  # not necessary when using dataframe from model output
  noNA <- df[complete.cases(df),]

  
  # predict phenotype - here, probability of mortality
  df$pred <- predict(mod, type = 'zprob', re.form = re.form, se.fit = se.fit)
  
  # un-log-transform actual height
  df$pheno <- exp(df$pheno) - 1
  
  # convert growth increment to binary 0/1
  # if value is not 0, convert to 1
  df$alive <- ifelse(df$pheno == 0, 0, 1)
  
  # logistic model
  mort <- glm(alive ~ pred, family = 'binomial', data = df)
  
  # this really should be a logistic fit not linear, but using this for now
  cor <- cor.test(df$alive, df$pred)
  
  return(list(actual = df$alive, predicted = df$pred, cor = cor, data_noNAs = noNA))
  
}

plot_predicted_vs_actual_mortality <- function(input, title = NULL, col = rgb(0,0,0,0.1), col_1to1 = 'red', col_fit = 'blue'){
  
  pval <- round(input$cor$p.value, 4)
  cor_est <- as.numeric(round(input$cor$estimate, 3))
  main <- paste(title, '\n', 'R = ', cor_est, ' | p = ', pval, sep = '')
  
  plot(input$actual, input$predicted, 
       pch = 16, 
       col = col, 
       xlab = "Mortality (actual)", 
       ylab = "Probability of mortality",
       title(main, adj = 0))
  #abline(0, 1, col = col_1to1, lty = 2) # 1to1 line
  abline(lm(input$predicted ~ input$actual), col = col_fit) # fit line
  
  # add text
  
 # legend('bottomright', fill = c(col_1to1, col_fit), legend = c('1:1 line', 'Best fit'))
  
}

# compare the distribution of residuals among multiple models
plot_resid <- function(mods, ...){
  
  
  # mods should be a named list of model outputs
  
  # set colors - cols25() with some similar colors removed
  cols <- c("#1F78C8", "#ff0000", "#33a02c", "#6A33C2", "#ff7f00", "#FFD700", "#a6cee3", "#FB6496", "#b2df8a", "#CAB2D6", "#FDBF6F", "#999999", "#EEE685", "#C8308C", "#FF83FA", "#C814FA", "#0000FF", "#36648B", "#00E2E5", "#00FF00", "#778B00", "#BEBE00", "#8B3B00", "#A52A3C")
  
  
  par(mfrow = c(1,3))
  
  # calculate residuals for all models
  res <- list()
  for(n in 1:length(mods)){
    res[[n]] <- residuals(mods[[n]])
  }
  
  ##############################
  # version with transformed residuals (from log scale to cm)
  res.cm <- list()
  for(n in 1:length(res)){
    res.cm[[n]] <- exp(res[[n]])-1
  }
  # plot first one
  # ugly code to get the max value for y axis
  ymax <- max(sapply(1:length(res.cm), FUN =  function(x) max(density(res.cm[[x]])$y)))
  plot(density(res.cm[[1]]), col = cols[1], ylim = c(0, ymax), main = 'residuals in cm')
  abline(v = mean(res.cm[[1]]), col = cols[1])
  # add the rest
  for(n in 2:length(mods)){
    # get residuals and un-log-transfrom
    lines(density(res.cm[[n]]), col = cols[n])
    abline(v = mean(res.cm[[n]]), col = cols[n])
  }
  
  abline(v = 0, lty = 2)
  legend('topright', 
         legend = names(mods),
         fill = cols[1:length(mods)])
  
  #####################################
  # untransformed (log scale) version
  
  # ugly code to get the max value for y axis
  ymax <- max(sapply(1:length(res), FUN =  function(x) max(density(res[[x]])$y)))
  
  # plot first one

  plot(density(res[[1]]), col = cols[1], main = 'log-transformed residuals', ylim = c(0,ymax))
  abline(v = mean(res.cm[[1]]), col = cols[1])
  # add the rest
  for(n in 2:length(mods)){
    lines(density(res[[n]]), col = cols[n])
    abline(v = mean(res.cm[[n]]), col = cols[n])
  }
  
  abline(v = 0, lty = 2)
  legend('topright', 
         legend = names(mods),
         fill = cols[1:length(mods)])
  
  ########################################
  # use absolute value to compare how "wrong" each model is without regard to over/underestimation
  
  # calc absolute values
  res.abs <- list()
  for(n in 1:length(res)){
    res.abs[[n]]<- abs(res[[n]])
  }
  
  # ugly code to get the max value for y axis
  ymax <- max(sapply(1:length(res.abs), FUN =  function(x) max(density(res.abs[[x]])$y)))
  
  # plot first one
  plot(density(res.abs[[1]]), col = cols[1], ylim = c(0, ymax), main = 'absolute log-transformed residuals')
  abline(v = mean(res.cm[[1]]), col = cols[1])
  # add the rest
  for(n in 2:length(mods)){
    lines(density(res.abs[[n]]), col = cols[n])
    abline(v = mean(res.cm[[n]]), col = cols[n])
  }
  
  abline(v = 0, lty = 2)
  legend('topright', 
         legend = names(mods),
         fill = cols[1:length(mods)])
  
  # reset par
  par(mfrow = c(1,1))
  
}

```


# Multi-year Model Setup

## Plot MCMT over three years

```{r plot_climate_variables}

# the MCMT for each year is highly correlated but shows yearly variation
par(mfrow = c(1,3))

plot(dat$garden_MCMT_2021, dat$garden_MCMT_2020, ylim = c(-16,15))
abline(0,1, col = 'blue')
plot(dat$garden_MCMT_2021, dat$garden_MCMT_2022, ylim = c(-16,15))
abline(0,1, col = 'blue')
plot(dat$garden_MCMT_2021, dat$garden_MCMT_2023, ylim = c(-16,15))
abline(0,1, col = 'blue')


```

## Select variables and merge years into one dataframe

```{r setup_run_glmTMB_all_years}

# select the climate variables being used and merge growth values from 2021 and 2022 into the same column, adding a random effect for year
# to run the model using a climate variable other than MCMT, change 'colname' variables here
# NOT including 2023 data in this script because the loss of extreme warm and cold gardens resulted in weird response curves (eg U-shaped curves with height increasing as temperatures became more extreme)
# note: some effects included here were used in testing models in previous versions, but are not used in the final model

# make a subset dataframe for 2021
# we need to include a random effect for the individual (to account for repeated sampling) and for year

# labels
pheno_label <- 'Growth Increment 2021-2022'
clim_label <- 'MCMT'

# for 2021
garden_clim_colname <- 'garden_MCMT_2021'
home_clim_colname <- 'provenance_MCMT'

# phenotype to use
pheno_colname <- 'GrowthIncrement_2021'

# setup variables

# set climate and phenotype variables
garden_clim <- dat[,garden_clim_colname]
garden_clim_2 <- dat[,garden_clim_colname]^2
home_clim <- dat[,home_clim_colname]
home_clim_2 <- dat[,home_clim_colname]^2
pheno <- dat[,pheno_colname]

# add other random and fixed effects
block <- as.character(interaction(dat$MiniCG_Site, dat$block, drop = T))
Pt <- dat$Pt
indiv <- as.character(dat$Unique_ID)
genotype <- as.character(dat$Genotype)
garden <- as.character(dat$MiniCG_Site)
pc1 <- dat$genetic_PC1
pc2 <- dat$genetic_PC2
pc3 <- dat$genetic_PC3
pc4 <- dat$genetic_PC4
pc5 <- dat$genetic_PC5

# put in df
df.21 <- data.frame(pheno, garden_clim, garden_clim_2, home_clim, home_clim_2, Pt, indiv, genotype, garden, block, pc1, pc2, pc3, pc4, pc5)
df.21$year <- '2021'
str(df.21)

###################################
# now do the same for 2022 

garden_clim_colname <- 'garden_MCMT_2022'
home_clim_colname <- 'provenance_MCMT'

# phenotype to use
pheno_colname <- 'GrowthIncrement_2022'


# setup variables

# set climate and phenotype variables
garden_clim <- dat[,garden_clim_colname]
garden_clim_2 <- dat[,garden_clim_colname]^2
home_clim <- dat[,home_clim_colname]
home_clim_2 <- dat[,home_clim_colname]^2
pheno <- dat[,pheno_colname]

# add other random and fixed effects
block <- as.character(interaction(dat$MiniCG_Site, dat$block, drop = T))
Pt <- dat$Pt
indiv <- as.character(dat$Unique_ID)
genotype <- as.character(dat$Genotype)
garden <- as.character(dat$MiniCG_Site)
pc1 <- dat$genetic_PC1
pc2 <- dat$genetic_PC2
pc3 <- dat$genetic_PC3
pc4 <- dat$genetic_PC4
pc5 <- dat$genetic_PC5

# put in df
df.22 <- data.frame(pheno, garden_clim, garden_clim_2, home_clim, home_clim_2, Pt, indiv, genotype, garden, block, pc1, pc2, pc3, pc4, pc5)
df.22$year <- '2022'
str(df.22)

# cleanup 2022
# remove the gardens that weren't measured in 2022
df.22 <- df.22[df.22$garden %in% levels(dat$MiniCG_Site_2022),]
# remove NWMO, which is missing preflush heights
df.22 <- df.22[! df.22$garden %in% 'NWMO',]
# remove trees that were dead in both 2021 and 2022
dead <- rownames(dat)[dat$Survival_09_2021 == 0 & dat$Survival_09_2022 == 0]
df.22 <- df.22[! df.22$indiv %in% dead,]


######################################
# combine 2021 and 2022
df <- merge(df.21, df.22, all = T )
str(df)

# add rownames and get list of alive individuals for each year
rownames(df) <- paste(df$indiv, df$year, sep = '_')


alive <- c(paste(alive21, '_2021', sep = ''),
           paste(alive22, '_2022', sep = ''))

dim(df)
dim(df[alive,])

#checks

# look at growth increment
par(mfrow = c(1,2))
hist(df$pheno)
hist(log(df$pheno + 1))

# variation across gardens/years
par(mfrow = c(1,1), mar = c(10,4,4,3))
boxplot(df$pheno~ df$year * df$garden, las = 2, xlab = '', col = c('lightblue', 'dodgerblue3'))

#check climate - make sure home/garden climate was assigned correctly
boxplot(df$garden_clim ~ df$year * df$garden, las = 2, xlab = '', col = c('lightblue', 'dodgerblue3'), drop = T)
boxplot(df$home_clim ~ df$year * df$garden, las = 2, xlab = '', col = c('lightblue', 'dodgerblue3'), drop = T)

boxplot(df$home_clim ~ df$genotype, las = 2, xlab = '')

# reset colnames - these get used for labeling outputs later
garden_clim_colname <- 'garden_MCMT_2021-2022'
home_clim_colname <- 'provenance_MCMT'
pheno_colname <- 'GrowthIncrement_2021-2022'

# check correlations among variables
pairs.panels(df, scale = T)

```


## Scale variables


```{r models_with_scaled_variables_setup}

# scale variables - having them on a similar scale improves model convergence. Previously got warnings about variables being on different scales.

dput(colnames(df))

# just scale numeric variables, then re-combine


# don't center - this means scale factor will be the same for each variable (column) so back-transforming is easier
orig <- df[,c("garden_clim", "garden_clim_2", "home_clim", "home_clim_2", "Pt", "pc1", "pc2", "pc3", "pc4", "pc5")]
scaled <- scale(orig, center = F)

# merge
df.scaled <- data.frame(df[, c('pheno', 'genotype', 'garden', 'block', 'indiv', 'year')], scaled)

# plot
pairs(df.scaled[,c("pheno", "garden_clim", "garden_clim_2", "home_clim", "home_clim_2", "Pt", "pc1", "pc2", "pc3", "pc4", "pc5")])

# get scale factors for back-transforming
summary(scaled/orig)

# get and save the first line (which is the same for all rows in a column)
# convert real data to scaled data: orig * scaling_factor
# convert scaled data to real data: scaled/scaling_factor
scaled_orig_df <- scaled/orig
scaling_factor <- as.vector(scaled_orig_df[1,])

```



# Run multiyear models

## Compare setting year as a fixed vs random effect

Fit a zero-inflated model predicting growth (conditional) and mortality (zero-inflated) for each individual based on garden MCMT, home MCMT, their square terms (accounting for parabolic response to temperature), genetic structure represented by genetic PCs 1-3, and interaction effects between genetic PCs and garden environment (GxE) and home climate and garden environment (GxE when home climate is a proxy for genetic variation due to adaptation).

This section compares setting year as a fixed effect vs random effect.

```{r compare_multiyear_models}


# info about repeated measures
# https://stackoverflow.com/questions/63360751/specifying-random-effects-for-repeated-measures-in-logistic-mixed-model-in-r-lm
# https://academic.oup.com/treephys/article/37/1/33/2527789#58152734


# set up list to save different models to for comparison
mods <- list()

# year as RE, individual as separate RE
# used this one for the final version!

mods$year_indiv_RE <- glmmTMB(log(pheno + 1) ~ garden_clim*home_clim + garden_clim_2*home_clim_2  + garden_clim_2 * home_clim + home_clim_2* garden_clim +  pc1*garden_clim + pc2*garden_clim + pc3*garden_clim + pc1*garden_clim_2 + pc2*garden_clim_2 + pc3*garden_clim_2 +
                      (1 | genotype) + (1 | garden/block) +(1 | year) + (1 | indiv),
               data = df.scaled,
               family = gaussian(),
               ziformula = ~.)

summary(mods$year_indiv_RE)

plot_model(mods$year_indiv_RE, type = 'std', vline.color = "black", show.values = T) +
   theme(text=element_text(size=16)) + 
  ggtitle('multiyear test')

plot_model(mods$year_indiv_RE, type = 're', 
                  sort.est = 'sort.all', 
                  grid = F, 
                  vline.color = 'black', 
                  wrap.title = 20, wrap.labels = 20)


# use genotype as  RE
# year as FE
mods$yearFE_indivRE <- glmmTMB(log(pheno + 1) ~ garden_clim*home_clim + garden_clim_2*home_clim_2  + garden_clim_2 * home_clim + home_clim_2* garden_clim 
                    + pc1 + pc2 + pc3 + pc1*garden_clim + pc2*garden_clim + pc3*garden_clim + pc1*garden_clim_2 + pc2*garden_clim_2 + pc3*garden_clim_2 + 
                      year +
                      (1 | genotype) + (1 | garden/block) + (1 | indiv),
               data = df.scaled, 
               family = gaussian(), 
               ziformula = ~.)


summary(mods$yearFE_indivRE)

plot_model(mods$yearFE_indivRE, type = 'std', vline.color = "black", show.values = T) +
   theme(text=element_text(size=16)) + 
  ggtitle('multiyear test')



# compare models
plot_models(mods, std.est = T, vline.color = "black", show.values = F, p.shape = T, m.labels = names(mods))

mods.yrs <- list()


par(mar = c(5,4,4,3), mfrow = c(2,2))
for(n in 1:length(mods)){
  
  mod <- mods[[n]]
  
  # overall model
  plot_predicted_vs_actual(model_R(mod, re.form = NULL), title = paste(names(mods)[n], 'overall, with random effects', sep = '\n'))
  plot_predicted_vs_actual(model_R(mod, re.form = NA), title = 'overall, no random effects')
  # conditional model (just growth)
  # with dead trees removed
  plot_predicted_vs_actual(model_R(mod, re.form = NULL, type = 'conditional'), title = 'conditional, with random effects')
  plot_predicted_vs_actual(model_R(mod, re.form = NA, type = 'conditional'), title = 'conditional, no random effects')
  
}

```


```{r save_outputs}

# which model to save?
mod <- mods$year_indiv_RE

save(mod, df.scaled, scaling_factor,
     file = paste('results/model_prediction/glmTMB_multiyear_model_outputs_', garden_clim_colname, '_vs_', pheno_colname, '.Rda', sep = ''))

```

# Plots for selected model

```{r model_plots}

# which model to use?
mod <- mods$year_indiv_RE

# print call
mod$call


# nice model plot using sjPlot
# see ?plot_model
# https://github.com/strengejacke/sjPlot/blob/6f57b80824575cc238834551be7270f8ecb19c8c/R/plot_type_est.R#L6

# get names of terms, excluding intercept
terms <- names(fixef(mod)[[1]][-1])
labs <- rep(NA, length = length(terms))
# add labels for the MCMT model
labs <- c("Garden MCMT", "Home MCMT", "Garden MCMT^2", "Home MCMT^2", "Genetic PC1", "Genetic PC2", "Genetic PC3", "Garden MCMT x Home MCMT", "Garden MCMT^2 x Home MCMT^2", "Garden MCMT^2 x Home MCMT", "Garden MCMT x Home MCMT^2", "Garden MCMT x Genetic PC1", "Garden MCMT x Genetic PC2", "Garden MCMT x Genetic PC3", "Garden MCMT^2 x Genetic PC1", "Garden MCMT^2 x Genetic PC2", "Garden MCMT^2 x Genetic PC3")
names(labs) <- terms

# check that they match
labs
cbind(labs, names(labs))

# reorder effect names 
ord <- c(1,3,2,4,5:17)

# png(file = paste('results/model_prediction/glmTMB_model_plot_standardized_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''),
# height = 12,
# width = 9,
# res = 300,
# units = 'in')

plot_model(mod, type = 'std', vline.color = "black", show.values = T, value.offset = 0.4, wrap.labels = 30, dot.size = 2) +
   theme(text=element_text(size=16)) +
  ggtitle('Annual Growth Increment (log transformed)') +
  scale_x_discrete(limits = rev(names(labs)[ord]), labels = labs[ord])

#dev.off()



# plot predicted vs actual height values
# info on including random effects: https://stats.stackexchange.com/questions/191648/using-re-form-in-predict-mermod-for-a-lmer-model


# png(file = paste('results/model_prediction/glmTMB_full_model_actual_vs_predicted_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''),
#     height = 10,
#     width = 10,
#     res = 300,
#     units = 'in')

par(mar = c(5,4,4,3), mfrow = c(2,2))
plot_predicted_vs_actual(model_R(mod, re.form = NULL), title = 'overall, with random effects')
plot_predicted_vs_actual(model_R(mod, re.form = NA), title = 'overall, no random effects')
plot_predicted_vs_actual(model_R(mod, re.form = NULL, type = 'conditional',), title = 'conditional, with random effects')
plot_predicted_vs_actual(model_R(mod, re.form = NA, type = 'conditional',), title = 'conditional, no random effects')


#dev.off()

# looking at model
# see ?Anova.glmmTMB
# anova: https://stats.stackexchange.com/questions/60362/choice-between-type-i-type-ii-or-type-iii-anova
summary(mod)


# create model table (using sjPlot package)
# on standardizing (from ?tab.model):
# "Default standardization is done by completely refitting the model on the standardized data. Hence, this approach is equal to standardizing the variables before fitting the model, which is particularly recommended for complex models that include interactions or transformations (e.g., polynomial or spline terms)."
tab_model(mod, show.std = T, show.aic = T)
# save
# tab_model(mod, show.std = T, show.aic = T,
#           file = paste('results/model_prediction/glmTMB_model_anova_table_', garden_clim_colname, '_vs_', pheno_colname, '.html', sep = ''))

# these produce the same results
# car::Anova(mod, component = "cond", type = 2)
# car::Anova(mod, component = "zi", type = 2)
glmmTMB:::Anova.glmmTMB(mod, component = 'cond', type = 2)
glmmTMB:::Anova.glmmTMB(mod, component = 'zi', type = 2)

# print fixed and random effects
fixef(mod)
#ranef(mod) # skipping this because the RE for each individual prints a lot of text 

# plot random effects
# plotting together doesn't save well, plot them all separately then combine later
res <- plot_model(mod, type = 're', 
                  sort.est = 'sort.all', 
                  grid = F, 
                  vline.color = 'black', 
                  wrap.title = 20, wrap.labels = 20)

plot(res[[3]]) +
  labs(tag = 'A')
#ggsave(paste('results/model_prediction/glmTMB_model_randomEffects_garden_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''),
#             height = 9, width = 8)

plot(res[[2]]) +
  labs(tag = 'B')
#ggsave(paste('results/model_prediction/glmTMB_model_randomEffects_block_garden_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''),
#            height = 9, width = 8)

plot(res[[1]]) +
  labs(tag = 'C')
#ggsave(paste('results/model_prediction/glmTMB_model_randomEffects_genotype_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''),
#             height = 9, width = 8)

plot(res[[4]]) +
  labs(tag = 'D')
#ggsave(paste('results/model_prediction/glmTMB_model_randomEffects_year_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''),
#            height = 9, width = 8)


# save individual random effects save as text - too big to easily visualize on plot
res <- ranef(mod)
#write.csv(res$cond$indiv, file = paste('results/model_prediction/glmTMB_model_randomEffects_conditional_individual_', garden_clim_colname, '_vs_', pheno_colname, '.csv', sep = ''))
#write.csv(res$zi$indiv, file = paste('results/model_prediction/glmTMB_model_randomEffects_zero-inflated_individual_', garden_clim_colname, '_vs_', pheno_colname, '.csv', sep = ''))


###############
# plots of residuals
# do residuals vary by home climate?
par(mfrow = c(1,1))
plot(mod$frame$home_clim[1:44], res$cond$genotype$`(Intercept)`)

# plots of residuals
# distribution
hist(residuals(mod))

# residuals for each effect
plot_model(mod, type = 'resid')
plot_residuals(mod, show.pred = F)
# residuals vs height
#plot(pheno[! is.na(pheno)], residuals(mod))
hist(residuals(mod))

# check model assumptions
# "Note: For mixed models, the diagnostic plots like linear relationship or check for Homoscedasticity, do not take the uncertainty of random effects into account, but is only based on the fixed effects part of the model."
# "For linear (mixed) models, plots for multicollinearity-check (Variance Inflation Factors), QQ-plots, checks for normal distribution of residuals and homoscedasticity (constant variance of residuals) are shown. For generalized linear mixed models, returns the QQ-plot for random effects."
plot_model(mod, type = 'diag')


# model fit 
sjp.poly(mod, poly.term = 'garden_clim', poly.degree = 2)

# visualize effects of factors on growth
plot_model(mod, type = 'pred', terms = 'pc1')
plot_model(mod, type = 'pred', terms = 'pc2')
plot_model(mod, type = 'pred', terms = 'pc3')
plot_model(mod, type = 'pred', terms = c('pc1', 'pc3'))
plot_model(mod, type = 'pred', terms = 'home_clim')
plot_model(mod, type = 'pred', terms = 'home_clim_2')
plot_model(mod, type = 'pred', terms = c('home_clim', 'garden_clim'))

# correlation of random effects from the conditional and zero-inflated model
par(mfrow = c(1,1))
plot(ranef(mod)$cond$garden[,1], ranef(mod)$zi$garden[,1])
plot(ranef(mod)$cond$genotype[,1], ranef(mod)$zi$genotype[,1])

# plot genetic PCs, colored by species ancestry

par(mfrow = c(1,2))
# label by transect
plot(dat$genetic_PC1, dat$genetic_PC3, type = 'n')
text(dat$genetic_PC1, dat$genetic_PC3, dat$transect, col = dat$color_Pt)
# label by genotyoe
plot(dat$genetic_PC1, dat$genetic_PC3, type = 'n')
text(dat$genetic_PC1, dat$genetic_PC3, dat$Genotype, col = dat$color_Pt)

##########################
# plot actual vs predicted height for different subsets of data 

# each genotype separately
# when random effects are included
# use se.fit = T to get standard errors
pred <- predict(mod, type = 'response', se.fit = F)
# un-log-transform
pred <- exp(pred) - 1


genos <- sort(unique(df.scaled$genotype))

# use only data with no NAs (lm removes this automatically)
noNA <- df.scaled[,c("pheno", "garden_clim", "garden_clim_2", "home_clim", "home_clim_2", "genotype", "garden", "year", "pc1", "pc2", "pc3")]
noNA <- noNA[complete.cases(noNA),]

par(mfrow = c(4,4))
genos_r <- rep(NA, length(genos))
names(genos_r) <- genos
genos_p <- rep(NA, length(genos))
names(genos_p) <- genos

for(n in 1:length(genos)){
  
  geno <- genos[n]
  
  keep <- which(noNA$genotype == geno)
  
  if(length(keep) > 0){
    
    cortest <- cor.test(noNA$pheno[keep], pred[keep])
    pval <- round(cortest$p.value, 4)
    Rval <- round(cortest$estimate, 3)
    
    #par(mfrow = c(1,2))
    plot(noNA$pheno[keep], pred[keep], pch = 16, col = rgb(0,0,0,0.4), 
         main = paste( 'genotype ', geno, '\n', 'R = ', Rval, '  |  pval = ', pval, sep = ''), 
         xlab = "Height (actual)", 
         ylab = "Height (predicted)")
    abline(0, 1, col = 'red', lty = 2) # 1to1 line
    abline(lm(pred[keep] ~ noNA$pheno[keep]), col = 'blue') # fit line
    
    genos_r[geno] <- Rval
    genos_p[geno] <- pval
    
  }  

}

par(mfrow = c(1,2))
hist(genos_r)
hist(genos_r^2)


# plot each garden separately

gards <- unique(df.scaled$garden)

par(mfrow = c(4,5))
gards_r <- rep(NA, length(gards))
names(gards_r) <- gards


for(n in 1:length(gards)){
  
  gard <- gards[n]
  
  # use only data with no NAs (lm removes this automatically)
  noNA <- df.scaled[,c("pheno", "garden_clim", "garden_clim_2", "home_clim", "home_clim_2", "genotype", "garden", "year", "pc1", "pc2", "pc3")]
  noNA <- noNA[complete.cases(noNA),]
  
  keep <- which(noNA$garden == gard)
  
  if(length(keep) > 0){
    
    cortest <- cor.test(noNA$pheno[keep], pred[keep])
    pval <- round(cortest$p.value, 4)
    Rval <- round(cortest$estimate, 3)
    
    #par(mfrow = c(1,2))
    plot(noNA$pheno[keep], pred[keep], pch = 16, col = rgb(0,0,0,0.4), 
         main = paste( 'garden ', gard, '\n', 'R = ', Rval, '  |  pval = ', pval, sep = ''), 
         xlab = "Height (actual)", 
         ylab = "Height (predicted)")
    abline(0, 1, col = 'red', lty = 2) # 1to1 line
    abline(lm(pred[keep] ~ noNA$pheno[keep]), col = 'blue') # fit line
    
    gards_r[gard] <- Rval
    
  }  

}

hist(gards_r)
hist(gards_r^2)
par(mfrow = c(1,1))
barplot(gards_r, las = 2)

# Each year separately

years <- unique(df.scaled$year)

par(mfrow = c(1,2))
years_r <- rep(NA, length(years))
names(years_r) <- years

for(n in 1:length(years)){
  
  year <- years[n]
  
  # use only data with no NAs (lm removes this automatically)
  noNA <- df.scaled[,c("pheno", "garden_clim", "garden_clim_2", "home_clim", "home_clim_2", "genotype", "garden", "year", "pc1", "pc2", "pc3")]
  noNA <- noNA[complete.cases(noNA),]
  
  keep <- which(noNA$year == year)
  
  if(length(keep) > 0){
    
    cortest <- cor.test(noNA$pheno[keep], pred[keep])
    pval <- round(cortest$p.value, 4)
    Rval <- round(cortest$estimate, 3)
    
    #par(mfrow = c(1,2))
    plot(noNA$pheno[keep], pred[keep], pch = 16, col = rgb(0,0,0,0.4), 
         main = paste( 'Year ', year, '\n', 'R = ', Rval, '  |  pval = ', pval, sep = ''), 
         xlab = "Height (actual)", 
         ylab = "Height (predicted)")
    abline(0, 1, col = 'red', lty = 2) # 1to1 line
    abline(lm(pred[keep] ~ noNA$pheno[keep]), col = 'blue') # fit line
    
    years_r[year] <- Rval
    
  }  

}

hist(years_r)
hist(years_r^2)
par(mfrow = c(1,1))
barplot(years_r, las = 2)

# each garden + year separately

gards <- unique(df.scaled$garden)

par(mfrow = c(4,5))
# gards_r <- rep(NA, length(gards))
# names(gards_r) <- gards

# nested loop - loop through gardens then years (some gardens not measured in some years)
for(n in 1:length(gards)){
  
  gard <- gards[n]
  
  years <- sort(unique(df.scaled[df.scaled$garden == gard, 'year']))
  
  for(y in 1:length(years)){
    
    year <- years[y]
    
    # use only data with no NAs (lm removes this automatically)
    noNA <- df.scaled[,c("pheno", "garden_clim", "garden_clim_2", "home_clim", "home_clim_2", "genotype", "garden", "year", "pc1", "pc2", "pc3")]
    noNA <- noNA[complete.cases(noNA),]
    
    keep <- which(noNA$garden == gard & noNA$year == year)
    
    if(length(keep) > 0){
      
      cortest <- cor.test(noNA$pheno[keep], pred[keep])
      pval <- round(cortest$p.value, 4)
      Rval <- round(cortest$estimate, 3)
      
      #par(mfrow = c(1,2))
      plot(noNA$pheno[keep], pred[keep], pch = 16, col = rgb(0,0,0,0.4), 
           main = paste(gard, ' ', year, '\n', 'R = ', Rval, '  |  pval = ', pval, sep = ''), 
           xlab = "Height (actual)", 
           ylab = "Height (predicted)")
      abline(0, 1, col = 'red', lty = 2) # 1to1 line
      
      # skip plotting fit line if it can't run the correlation (eg SWMN 2022 are all dead)
      # I haven't gotten this to work
      #try(x <- abline(lm(pred[keep] ~ noNA$pheno[keep]), col = 'blue'), silent = T)
      
      # gards_r[gard] <- Rval
      
    }  
  }
}


# does ancestry affect how well the model predicts height?
# might expect hybrids to be less predictable
r_vs_pt <- data.frame(cor_R = genos_r, cor_pval = genos_p,  Pt = NA, hybrid_index = NA)

r_vs_pt 


# get Pt (tricho ancestry)
for(n in 1:nrow(r_vs_pt)){
  
  r_vs_pt[n, 'Pt'] <- dat[match(rownames(r_vs_pt)[n], df$genotype), 'Pt']
  r_vs_pt[n, 'hybrid_index'] <- dat[match(rownames(r_vs_pt)[n], df$genotype), 'hybrid_index_Pb']
  
}
par(mfrow = c(1,2))
plot(r_vs_pt$Pt, r_vs_pt$cor_R)
plot(r_vs_pt$hybrid_index, r_vs_pt$cor_R)

```

# Model validation using subsets

Test model performance using leave-one-out cross-validation: Split data into training/test sets by leaving one garden out, training the model without that garden, then predict for that garden.

## By garden

This section separates models by garden, but not by year - so both years are included in the plots of actual vs predicted values. The MS does not report values from this section, instead reporting years separately as in the next code chunk.

```{r validation_leave_garden_out}


# validate model by removing one garden to build model, then testing predictions for that garden

# plot full model for comparison
plot_model(mod, type = 'std', vline.color = "black", show.values = T) +
   theme(text=element_text(size=16)) + 
  ggtitle('Full model')


# loop though gardens and leave each out
# not all of these models converge

gards <- sort(unique(df.scaled$garden))

mods_test <- list()

for(n in 1:length(gards)){
  
  gard <- gards[n]
  
  # subset
  df.scaled.sub <- df.scaled[df.scaled$garden != gard,]
  
  mods_test[[n]] <- update(mod, data = df.scaled.sub)
  
  summary(mods_test[[n]])
  
  p <- plot_model(mods_test[[n]], type = 'std', vline.color = "black", show.values = T) +
    theme(text=element_text(size=16)) + 
    ggtitle(paste('without', gard))
  plot(p)
  
  
}

names(mods_test) <- paste('without_', gards, sep = '')

plot_models(mods_test, std.est = 'std', spacing = 0.8, vline.color = "black", show.values = F, p.shape = T, m.labels = names(mods_test), colors = c("#008856",  "#8DB600", "#BE0032", "#875692", "#A1CAF1", "#2B3D26", "#C2B280", "#F3C300", "#848482", "#0067A5", "#E68FAC", "#F99379", "#604E97", "#B3446C", "#DCD300", "#882D17", "#E25822", "#654522"))


#ggsave(filename = paste('results/model_prediction/compare_models_leave_one_garden_out_', garden_clim_colname, '_vs_', pheno_colname, '.pdf', sep = ''), device = 'pdf', height = 12, width = 8)

########

# residuals for each sub-model
#pdf(file = 'results/model_prediction/residuals_density_leave_garden_out.pdf', height = 6, width = 16)
#plot_resid(mods_test)
#dev.off()

#########
# plot predicted vs actual for the full dataset (all gardens) for each model

par(mfrow = c(2,2))

for(n in 1:length(mods_test)){
  
  modN <- mods_test[[n]]
  
      # overall model
    plot_predicted_vs_actual(model_R(modN, re.form = NULL), title = names(mods_test)[n])
    plot_predicted_vs_actual(model_R(modN, re.form = NA), title = 'no random effects')
    # conditional model (just growth)
    # with dead trees removed
    plot_predicted_vs_actual(model_R(modN, re.form = NULL, type = 'conditional'), title = 'conditional')
    plot_predicted_vs_actual(model_R(modN, re.form = NA, type = 'conditional'), title = 'no random effects')
  
}



##########
# predict for each garden left out of model


# function for predicting with new data - based on model_R() function defined above
model_R_new <- function(mod, newdata, type = 'response', re.form = NULL, se.fit = FALSE){
  
  df <- newdata
  
  # make new dataframe with no NA values
  noNA <- df[complete.cases(df),]
  
  # predict phenotype (generally height)
  noNA$pred <- predict(mod, newdata = noNA, type = type, re.form = re.form, se.fit = se.fit, allow.new.levels=TRUE)
  
  # un-log-transform
  noNA$pred <- exp(noNA$pred) - 1
  
  # if type = conditional, remove dead trees (which are modeled separately in zero-inflated model)
  if(type == 'conditional'){
    noNA <- noNA[rownames(noNA) %in% alive,]
  }
  
  # calculate correlation
  cor <- cor.test(noNA$pheno, noNA$pred)
  
  return(list(actual = noNA$pheno, predicted = noNA$pred, cor = cor, data_noNAs = noNA))
  
}


# loop through each model and predict for garden that was left out

# objects to save results to
pred.overall_REs <- list()
pred.overall_noREs <- list()
pred.cond_REs <- list()
pred.cond_noREs <- list()

par(mfrow = c(2,2))
for(n in 1:length(gards)){
  
  gard <- gards[n]
  modN <- mods_test[[n]]
  
  newdata <- df.scaled[df.scaled$garden == gard,]
  
  # plot
  # overall model
  
  pred.overall_REs[[n]] <- model_R_new(modN, newdata =  newdata, re.form = NULL)
  plot_predicted_vs_actual(pred.overall_REs[[n]], title = paste(names(mods_test)[n], 'overall, with random effects', sep = '\n'))
  
  pred.overall_noREs[[n]] <- model_R_new(modN, newdata = newdata, re.form = NA)
  plot_predicted_vs_actual(pred.overall_noREs[[n]], title = 'overall, no random effects')
  
  
  # conditional model
  pred.cond_REs[[n]] <- model_R_new(modN, type = 'conditional', newdata =  newdata, re.form = NULL)
  plot_predicted_vs_actual(pred.cond_REs[[n]], title = 'conditional, with random effects')
  
  pred.cond_noREs[[n]] <- model_R_new(modN, type = 'conditional', newdata = newdata, re.form = NA)
  plot_predicted_vs_actual(pred.cond_noREs[[n]], title = 'conditional, no random effects')
  
  
}

# remove "all" model from names
model_names <- names(mods_test)[names(mods_test) != 'all']
names(pred.overall_REs) <- model_names
names(pred.overall_noREs) <- model_names
names(pred.cond_REs) <- model_names
names(pred.cond_noREs) <- model_names

preds.full <- list(pred.overall_REs, pred.overall_noREs, pred.cond_REs, pred.cond_noREs)
names(preds.full) <- c('overall_REs', 'overall_noREs', 'cond_REs', 'cond_noREs')

# plot panel with just the conditional RE plots
gard_cols <- c("#008856",  "#8DB600", "#BE0032", "#875692", "#A1CAF1", "#2B3D26", "#C2B280", "#F3C300", "#848482", "#0067A5", "#E68FAC", "#F99379", "#604E97", "#B3446C", "#DCD300", "#882D17", "#E25822", "#654522")


#png(filename = paste('results/model_prediction/validation_leaveOneGardenOutAndPredict_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''), height = 7.2, width = 12.8, units = 'in', res = 300)

#pdf(file = paste('results/model_prediction/validation_leaveOneGardenOutAndPredict_', garden_clim_colname, '_vs_', pheno_colname, '.pdf', sep = ''), height = 8, width = 12)
par(mfrow = c(3,4))

for(n in 1:length(pred.cond_noREs)){
  
  gard <- gsub('without_', '', names(pred.cond_noREs)[n])
  plot_predicted_vs_actual(pred.cond_noREs[[n]], 
                           title = gard, 
                           col = gard_cols[n],
                           col_1to1 = 'grey',
                           col_fit = gard_cols[n])
  
}

#dev.off()

# version ordered by mortality - use mean to get a proportion
# use as character so it's alphabetical to match 'gard' object
dead <- aggregate(dat$Survival_09_2021, by = list(as.character(dat$MiniCG_Site)), FUN = mean, na.rm = T)

ord <- order(dead$x, decreasing = T)

# proportion surviving in each garden (NOT proportion dead)
dead[ord,]

gard.sort <- dead$Group.1[ord]

#pdf(file = paste('results/model_prediction/validation_leaveOneGardenOutAndPredict_sortGardensByMortality', garden_clim_colname, '_vs_', pheno_colname, '.pdf', sep = ''), height = 8, width = 12)

par(mfrow = c(3,4))

for(n in ord){
  
  
  gard <- gards[n]
  
  plot_predicted_vs_actual(pred.cond_noREs[[n]], 
                           title = gard, 
                           col = gard_cols[n],
                           col_1to1 = 'grey',
                           col_fit = gard_cols[n])
  
}


#dev.off()

# png version for supplement

#png(filename = paste('results/model_prediction/validation_leaveOneGardenOutAndPredict_sortGardensByMortality', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''), height = 7.2, width = 12.8, units = 'in', res = 300)

par(mfrow = c(3,6))

for(n in ord){
  
  
  gard <- gards[n]
  
  plot_predicted_vs_actual(pred.cond_REs[[n]], 
                           title = gard, 
                           col = gard_cols[n],
                           col_1to1 = 'grey',
                           col_fit = gard_cols[n],
                           legend = F)
  
}


#dev.off()

# get all the R values
rvals <- as.data.frame(matrix(nrow = length(preds.full$overall_REs), ncol = 4))
colnames(rvals) <- names(preds.full)
rownames(rvals) <- names(preds.full$overall_REs)

rvals$overall_REs <- as.numeric(lapply(1:length(preds.full$overall_REs), function(x) preds.full$overall_REs[[x]]$cor$estimate))
rvals$overall_noREs <- as.numeric(lapply(1:length(preds.full$overall_noREs), function(x) preds.full$overall_noREs[[x]]$cor$estimate))
rvals$cond_REs <- as.numeric(lapply(1:length(preds.full$cond_REs), function(x) preds.full$cond_REs[[x]]$cor$estimate))
rvals$cond_noREs <- as.numeric(lapply(1:length(preds.full$cond_noREs), function(x) preds.full$cond_noREs[[x]]$cor$estimate))

rvals

# plots
par(mfrow = c(2,2))
# xlim assumes no negative correlations
hist(rvals$overall_REs, xlim = c(0,1))
hist(rvals$overall_noREs, xlim = c(0,1))
hist(rvals$cond_REs, xlim = c(0,1))
hist(rvals$cond_noREs, xlim = c(0,1))

colMeans(rvals)

rvals[which.max(rvals$cond_REs),]
rvals[which.min(rvals$cond_REs),]


```


## By garden + year

Predictions for each garden are made separately for each year, if multiple years of data are available (some gardens were only measured the first year). These are the results reported in the manuscript.

```{r validation_leave_garden_year_out}

# loop through each model and predict for garden that was left out
# here, predictions are separated by year 
# so model is trained without a garden, then predictions for that garden are made separately for each year
# use mods_test object from above chunk, with each garden removed 

gard_cols <- c("#008856",  "#8DB600", "#BE0032", "#875692", "#A1CAF1", "#2B3D26", "#C2B280", "#F3C300", "#848482", "#0067A5", "#E68FAC", "#F99379", "#604E97", "#B3446C", "#DCD300", "#882D17", "#E25822", "#654522")

# objects to save results to
pred.overall_REs <- list()
pred.overall_noREs <- list()
pred.cond_REs <- list()
pred.cond_noREs <- list()

gards <- sort(unique(df.scaled$garden))
cols.yr <- vector()

# check that garden names and model names are in the same order
cbind(gards, names(mods_test))

par(mfrow = c(2,2))
# counter to keep track of which garden/year we're on
count <- 1
for(n in 1:length(gards)){
  
  # loop through each garden, use model trained without that garden
  gard <- gards[n]
  modN <- mods_test[[n]]
  
  # subset data to this garden
  df.scaled.sub <- df.scaled[df.scaled$garden == gard,]
  
  # now loop through each year for this garden
  years <- sort(unique(df.scaled.sub$year))
  
  for(y in 1:length(years)){
    
    year <- years[y]
    
    # subset data to this year
    newdata <- df.scaled.sub[df.scaled.sub$year == year,]
    
    # plot actual vs predicted for this garden and year, using model trained without this garden
    
    # plots
    # overall model
    
    pred.overall_REs[[count]] <- model_R_new(modN, newdata =  newdata, re.form = NULL)
    plot_predicted_vs_actual(pred.overall_REs[[count]], title = paste(names(mods_test)[n], ', predicted for ', year, '\n', 'overall, with random effects', sep = ''))
    
    pred.overall_noREs[[count]] <- model_R_new(modN, newdata = newdata, re.form = NA)
    plot_predicted_vs_actual(pred.overall_noREs[[count]], title = 'overall, no random effects')
    
    
    # conditional model
    pred.cond_REs[[count]] <- model_R_new(modN, type = 'conditional', newdata =  newdata, re.form = NULL)
    plot_predicted_vs_actual(pred.cond_REs[[count]], title = 'conditional, with random effects')
    
    pred.cond_noREs[[count]] <- model_R_new(modN, type = 'conditional', newdata = newdata, re.form = NA)
    plot_predicted_vs_actual(pred.cond_noREs[[count]], title = 'conditional, no random effects')
    
    # name output
    name <- paste(gard, year, sep = '_')
    names(pred.overall_REs)[[count]] <- name
    names(pred.overall_noREs)[[count]] <- name
    names(pred.cond_REs)[[count]] <- name
    names(pred.cond_noREs)[[count]] <- name
    
    # set garden color and save to a vector
    cols.yr[count] <- gard_cols[n]
    
    # set counter
    count <- count + 1
  }
}


preds.full <- list(pred.overall_REs, pred.overall_noREs, pred.cond_REs, pred.cond_noREs)
names(preds.full) <- c('overall_REs', 'overall_noREs', 'cond_REs', 'cond_noREs')

# get yearly averages
pred.21 <- pred.cond_REs[grep('2021', names(pred.cond_REs))]
mean(sapply(1:length(pred.21), function(x) mean(pred.21[[x]]$cor$estimate)))

pred.22 <- pred.cond_REs[grep('2022', names(pred.cond_REs))]
mean(sapply(1:length(pred.22), function(x) mean(pred.22[[x]]$cor$estimate)))

# plot panel with just the conditional RE plots


#pdf(file = paste('results/model_prediction/validation_leaveOneGardenOutAndPredictYear_', garden_clim_colname, '_vs_', pheno_colname, '.pdf', sep = ''), height = 8, width = 12)
par(mfrow = c(3,4))

for(n in 1:length(pred.cond_REs)){
  
  
  title <- names(pred.cond_REs)[n]
  plot_predicted_vs_actual(pred.cond_REs[[n]], 
                           title = title, 
                           col = cols.yr[n],
                           col_1to1 = 'grey',
                           col_fit = cols.yr[n],
                           legend = F)
  
}

#dev.off()

#png(filename = paste('results/model_prediction/validation_leaveOneGardenOutAndPredictYear_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''), height = 16, width = 9.6, units = 'in', res = 300)

#par(mfrow = c(8,4))
par(mfrow = c(4,5))

for(n in 1:length(pred.cond_REs)){
  
  
  title <- names(pred.cond_REs)[n]
  plot_predicted_vs_actual(pred.cond_REs[[n]], 
                           title = title, 
                           col = cols.yr[n],
                           col_1to1 = 'grey',
                           col_fit = cols.yr[n],
                           legend = F)
  
}

#dev.off()

# get all the R values
#hist(as.numeric(lapply(1:length(preds.full$cond_REs), function(x) preds.full$cond_REs[[x]]$cor$estimate)))



```

# Compare model predictive ability when genetic and home climate information is removed

Test what information is needed to predict each genotype's response to temperature. if home climate can serve as a proxy for locally adaptive genetic variation, including genetic data may not be necessary for predicting growth under future climates.

```{r compare_noClimate_noGenetics}

# Does genetics or home climate give better predictions of fitness?
# using same method as above, calculate R^2 values when leaving each garden out
# but compare full model to models with home climate or genetic structure removed

# only climate, no genetics

gards <- unique(df.scaled$garden)

mods_test_no_gen <- list()

for(n in 1:length(gards)){
  
  gard <- gards[n]
  
  # subset
  df.scaled.sub <- df.scaled[df.scaled$garden != gard,]
  
  # run model
  mods_test_no_gen[[n]] <-  glmmTMB(log(pheno + 1) ~ garden_clim*home_clim + garden_clim_2*home_clim_2  + garden_clim_2*home_clim + home_clim_2*garden_clim +
                                      (1 | genotype) + (1 | garden/block) + (1 | year) + (1 | indiv),
                                    data = df.scaled.sub, 
                                    family = gaussian(), 
                                    ziformula = ~.)
  
summary(mods_test_no_gen[[n]])

p <- plot_model(mods_test_no_gen[[n]], type = 'std', vline.color = "black", show.values = T) +
  theme(text=element_text(size=16)) + 
  ggtitle(paste('without', gard))
plot(p)
  
  
}

names(mods_test_no_gen) <- paste('without_', gards, sep = '')

# plot together

plot_models(mods_test_no_gen, std.est = 'std', spacing = 0.8, vline.color = "black", show.values = F, p.shape = T, m.labels = names(mods_test), colors = c("#008856",  "#8DB600", "#BE0032", "#875692", "#A1CAF1", "#2B3D26", "#C2B280", "#F3C300", "#848482", "#0067A5", "#E68FAC", "#F99379", "#604E97", "#B3446C", "#DCD300", "#882D17", "#E25822", "#654522"))

#ggsave(filename = paste('results/model_prediction/compare_models_leave_one_garden_out_NO_GENETICS', garden_clim_colname, '_vs_', pheno_colname, '.pdf', sep = ''), device = 'pdf', height = 12, width = 8)


# loop through each model and predict for garden that was left out

# objects to save results to
pred.overall_REs <- list()
pred.overall_noREs <- list()
pred.cond_REs <- list()
pred.cond_noREs <- list()

par(mfrow = c(2,2))

# counter to keep track of which garden/year we're on
count <- 1
for(n in 1:length(gards)){
  
  # loop through each garden, use model trained without that garden
  gard <- gards[n]
  modN <- mods_test_no_gen[[n]]
  
  # subset data to this garden
  df.scaled.sub <- df.scaled[df.scaled$garden == gard,]
  
  # now loop through each year for this garden
  years <- sort(unique(df.scaled.sub$year))
  
  for(y in 1:length(years)){
    
    year <- years[y]
    
    # subset data to this year
    newdata <- df.scaled.sub[df.scaled.sub$year == year,]
    
    # plot actual vs predicted for this garden and year, using model trained without this garden
    
    # plots
    # overall model
    
    pred.overall_REs[[count]] <- model_R_new(modN, newdata =  newdata, re.form = NULL)
    plot_predicted_vs_actual(pred.overall_REs[[count]], title = paste(names(mods_test)[n], ', predicted for ', year, '\n', 'overall, with random effects', sep = ''))
    
    pred.overall_noREs[[count]] <- model_R_new(modN, newdata = newdata, re.form = NA)
    plot_predicted_vs_actual(pred.overall_noREs[[count]], title = 'overall, no random effects')
    
    
    # conditional model
    pred.cond_REs[[count]] <- model_R_new(modN, type = 'conditional', newdata =  newdata, re.form = NULL)
    plot_predicted_vs_actual(pred.cond_REs[[count]], title = 'conditional, with random effects')
    
    pred.cond_noREs[[count]] <- model_R_new(modN, type = 'conditional', newdata = newdata, re.form = NA)
    plot_predicted_vs_actual(pred.cond_noREs[[count]], title = 'conditional, no random effects')
    
    # name output
    name <- paste(gard, year, sep = '_')
    names(pred.overall_REs)[[count]] <- name
    names(pred.overall_noREs)[[count]] <- name
    names(pred.cond_REs)[[count]] <- name
    names(pred.cond_noREs)[[count]] <- name
    
    # set garden color and save to a vector
    #cols.yr[count] <- gard_cols[n]
    
    # set counter
    count <- count + 1
  }
}

preds.nogen <- list(pred.overall_REs, pred.overall_noREs, pred.cond_REs, pred.cond_noREs)
names(preds.nogen) <- c('overall_REs', 'overall_noREs', 'cond_REs', 'cond_noREs')



# only genetics and garden climate, not home climate

gards <- unique(df.scaled$garden)

mods_test_no_clim <- list()

for(n in 1:length(gards)){
  
  gard <- gards[n]
  
  # subset
  df.scaled.sub <- df.scaled[df.scaled$garden != gard,]
  
  # run model
  mods_test_no_clim[[n]] <-  glmmTMB(log(pheno + 1) ~ garden_clim + garden_clim_2  +
                                       pc1 + pc2 + pc3 + 
                                       pc1*garden_clim + pc2*garden_clim + pc3*garden_clim + pc1*garden_clim_2 + pc2*garden_clim_2 + pc3*garden_clim_2 +
                                       (1 | genotype) + (1 | garden/block) + (1 | year) + (1 | indiv),
                                     data = df.scaled.sub, 
                                     family = gaussian(), 
                                     ziformula = ~.)
  
  summary(mods_test_no_clim[[n]])
  
  p <- plot_model(mods_test_no_clim[[n]], type = 'std', vline.color = "black", show.values = T) +
    theme(text=element_text(size=16)) + 
    ggtitle(paste('without', gard))
  plot(p)
  
  
}

names(mods_test_no_clim) <- paste('without_', gards, sep = '')

# plot together

plot_models(mods_test_no_clim, std.est = 'std', spacing = 0.8, vline.color = "black", show.values = F, p.shape = T, m.labels = names(mods_test), colors = c("#008856",  "#8DB600", "#BE0032", "#875692", "#A1CAF1", "#2B3D26", "#C2B280", "#F3C300", "#848482", "#0067A5", "#E68FAC", "#F99379", "#604E97", "#B3446C", "#DCD300", "#882D17", "#E25822", "#654522"))

#ggsave(filename = paste('results/model_prediction/compare_models_leave_one_garden_out_NO_HOME_CLIMATE', garden_clim_colname, '_vs_', pheno_colname, '.pdf', sep = ''), device = 'pdf', height = 12, width = 8)


# loop through each model and predict for garden that was left out
# this uses the full model, no climate, and no genetics so they can be compared for each run without a garden

# objects to save results to
pred.overall_REs <- list()
pred.overall_noREs <- list()
pred.cond_REs <- list()
pred.cond_noREs <- list()

par(mfrow = c(2,2))
# counter to keep track of which garden/year we're on
count <- 1
for(n in 1:length(gards)){
  
  # loop through each garden, use model trained without that garden
  gard <- gards[n]
  modN <- mods_test_no_clim[[n]]
  
  # subset data to this garden
  df.scaled.sub <- df.scaled[df.scaled$garden == gard,]
  
  # now loop through each year for this garden
  years <- sort(unique(df.scaled.sub$year))
  
  for(y in 1:length(years)){
    
    year <- years[y]
    
    # subset data to this year
    newdata <- df.scaled.sub[df.scaled.sub$year == year,]
    
    # plot actual vs predicted for this garden and year, using model trained without this garden
    
    # plots
    # overall model
    
    pred.overall_REs[[count]] <- model_R_new(modN, newdata =  newdata, re.form = NULL)
    plot_predicted_vs_actual(pred.overall_REs[[count]], title = paste(names(mods_test)[n], ', predicted for ', year, '\n', 'overall, with random effects', sep = ''))
    
    pred.overall_noREs[[count]] <- model_R_new(modN, newdata = newdata, re.form = NA)
    plot_predicted_vs_actual(pred.overall_noREs[[count]], title = 'overall, no random effects')
    
    
    # conditional model
    pred.cond_REs[[count]] <- model_R_new(modN, type = 'conditional', newdata =  newdata, re.form = NULL)
    plot_predicted_vs_actual(pred.cond_REs[[count]], title = 'conditional, with random effects')
    
    pred.cond_noREs[[count]] <- model_R_new(modN, type = 'conditional', newdata = newdata, re.form = NA)
    plot_predicted_vs_actual(pred.cond_noREs[[count]], title = 'conditional, no random effects')
    
    # name output
    name <- paste(gard, year, sep = '_')
    names(pred.overall_REs)[[count]] <- name
    names(pred.overall_noREs)[[count]] <- name
    names(pred.cond_REs)[[count]] <- name
    names(pred.cond_noREs)[[count]] <- name
    
    # set garden color and save to a vector
    cols.yr[count] <- gard_cols[n]
    
    # set counter
    count <- count + 1
  }
}

preds.noclim <- list(pred.overall_REs, pred.overall_noREs, pred.cond_REs, pred.cond_noREs)
names(preds.noclim) <- c('overall_REs', 'overall_noREs', 'cond_REs', 'cond_noREs')

###########################
# compare R values of predictions among all three models

par(mfrow = c(1,3))
hist(as.numeric(lapply(1:length(preds.full$cond_REs), function(x) preds.full$cond_REs[[x]]$cor$estimate)))
hist(as.numeric(lapply(1:length(preds.nogen$cond_REs), function(x) preds.nogen$cond_REs[[x]]$cor$estimate)))
hist(as.numeric(lapply(1:length(preds.noclim$cond_REs), function(x) preds.noclim$cond_REs[[x]]$cor$estimate)))


# put all R values into a dataframe, each row being a model with one garden excluded, a column for full/conditional models and with/without random effects
# I did this in a pretty ugly way, oh well

mods_test_R <- data.frame(matrix(nrow = length(preds.full$overall_noREs)*3, ncol = 1))

colnames(mods_test_R) <- 'model'

mods_test_R$model <- c(
  rep('full', length(preds.full$cond_REs)),
  rep('no_genetics', length(preds.nogen$cond_REs)),
  rep('no_climate', length(preds.noclim$cond_REs))
)

mods_test_R$test_data <- c(
  names(preds.full$cond_REs),
  names(preds.nogen$cond_REs),
  names(preds.noclim$cond_REs)
)

# add overall with REs
mods_test_R$Rval_overall_REs <- c(
  as.numeric(lapply(1:length(preds.full$overall_REs), function(x) preds.full$overall_REs[[x]]$cor$estimate)),
  as.numeric(lapply(1:length(preds.nogen$overall_REs), function(x) preds.nogen$overall_REs[[x]]$cor$estimate)),
  as.numeric(lapply(1:length(preds.noclim$overall_REs), function(x) preds.noclim$overall_REs[[x]]$cor$estimate))
)

# add overall without REs
mods_test_R$Rval_overall_noREs <- c(
  as.numeric(lapply(1:length(preds.full$overall_noREs), function(x) preds.full$overall_noREs[[x]]$cor$estimate)),
  as.numeric(lapply(1:length(preds.nogen$overall_noREs), function(x) preds.nogen$overall_noREs[[x]]$cor$estimate)),
  as.numeric(lapply(1:length(preds.noclim$overall_noREs), function(x) preds.noclim$overall_noREs[[x]]$cor$estimate))
)

# add conditional with REs
mods_test_R$Rval_cond_REs <- c(
  as.numeric(lapply(1:length(preds.full$cond_REs), function(x) preds.full$cond_REs[[x]]$cor$estimate)),
  as.numeric(lapply(1:length(preds.nogen$cond_REs), function(x) preds.nogen$cond_REs[[x]]$cor$estimate)),
  as.numeric(lapply(1:length(preds.noclim$cond_REs), function(x) preds.noclim$cond_REs[[x]]$cor$estimate))
)

# add conditional no REs
mods_test_R$Rval_cond_noREs <- c(
  as.numeric(lapply(1:length(preds.full$cond_noREs), function(x) preds.full$cond_noREs[[x]]$cor$estimate)),
  as.numeric(lapply(1:length(preds.nogen$cond_noREs), function(x) preds.nogen$cond_noREs[[x]]$cor$estimate)),
  as.numeric(lapply(1:length(preds.noclim$cond_noREs), function(x) preds.noclim$cond_noREs[[x]]$cor$estimate))
)

# add garden and year information for plotting
mods_test_R$garden <-  sapply(1:nrow(mods_test_R), FUN = function(x) strsplit(mods_test_R$test_data, '_')[[x]][1])
mods_test_R$year <-  sapply(1:nrow(mods_test_R), FUN = function(x) strsplit(mods_test_R$test_data, '_')[[x]][2])

  
# calculate averages for each type of R calculated for the full model
colMeans(mods_test_R[mods_test_R$model == 'full',3:6])

# how many have accuracy >0.5?
sum(mods_test_R$Rval_cond_noREs[mods_test_R$model == 'full'] >0.5)

# ggplot settings
theme_set(theme_bw(base_size = 18))

gard_cols <- c("#008856",  "#8DB600", "#BE0032", "#875692", "#A1CAF1", "#2B3D26", "#C2B280", "#F3C300", "#848482", "#0067A5", "#E68FAC", "#F99379", "#604E97", "#B3446C", "#DCD300", "#882D17", "#E25822", "#654522")

# use ggpubr to add pvalues:
#https://www.r-bloggers.com/2017/06/add-p-values-and-significance-levels-to-ggplots/
my_comparisons <- combn(unique(mods_test_R$model), 2, simplify = F)

p1 <- ggplot(mods_test_R, aes(x = model, y = Rval_overall_REs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('r (Testing dataset)') +
  ggtitle('Overall model, with random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
plot(p1)

p2 <- 
  ggplot(mods_test_R, aes(x = model, y = Rval_overall_noREs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('r (Testing dataset)') +
  ggtitle('Overall model, no random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
  
plot(p2)
  
  
p3 <- ggplot(mods_test_R, aes(x = model, y = Rval_cond_REs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('r (Testing dataset)') +
  ggtitle('Conditional model, with random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
  
plot(p3)

p4 <- ggplot(mods_test_R, aes(x = model, y = Rval_cond_noREs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('r (Testing dataset)') +
  ggtitle('Conditional model, no random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
plot(p4)


wrap_plots(p1, p2, p3, p4)
# ggsave(filename = paste('results/model_prediction/Rvalues_predictWithoutOneGarden_compareClimateGeneticsModels_', garden_clim_colname, '_vs_', pheno_colname, '.pdf', sep = ''), device = 'pdf', height = 14, width = 16)
# ggsave(filename = paste('results/model_prediction/Rvalues_predictWithoutOneGarden_compareClimateGeneticsModels_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''), device = 'png', height = 14, width = 16)

# just the cond_noREs plot
ggplot(mods_test_R, aes(x = model, y = Rval_cond_noREs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('r (Testing dataset)') +
  ggtitle('Conditional model, no random effects') +
stat_compare_means(comparisons = my_comparisons, label = 'p.signif', method = 'wilcox.test', paired = T)
# ggsave(filename = paste('results/model_prediction/Rvalues_condNoREs_predictWithoutOneGarden_compareClimateGeneticsModels_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''), device = 'png', height = 6, width = 8)


##########
# just compare 2021

my_comparisons <- combn(unique(mods_test_R$model), 2, simplify = F)

p1 <- ggplot(subset(mods_test_R, year == '2021'), aes(x = model, y = Rval_overall_REs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('r (Testing dataset)') +
  ggtitle('Overall model, with random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
plot(p1)

p2 <- 
  ggplot(subset(mods_test_R, year == '2021'), aes(x = model, y = Rval_overall_noREs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('r (Testing dataset)') +
  ggtitle('Overall model, no random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
  
plot(p2)
  
  
p3 <- ggplot(subset(mods_test_R, year == '2021'), aes(x = model, y = Rval_cond_REs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('r (Testing dataset)') +
  ggtitle('Conditional model, with random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
  
plot(p3)

p4 <- ggplot(subset(mods_test_R, year == '2021'), aes(x = model, y = Rval_cond_noREs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('r (Testing dataset)') +
  ggtitle('Conditional model, no random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
plot(p4)


wrap_plots(p1, p2, p3, p4)
#ggsave(filename = paste('results/model_prediction/Rvalues_predictWithoutOneGarden_compareClimateGeneticsModels2021_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''), device = 'png', height = 14, width = 16)

#####################
# just compare 2022

my_comparisons <- combn(unique(mods_test_R$model), 2, simplify = F)

p1 <- ggplot(subset(mods_test_R, year == '2022'), aes(x = model, y = Rval_overall_REs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.1) +
  ylab('R (Testing dataset)') +
  ggtitle('Overall model, with random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
plot(p1)

p2 <- 
  ggplot(subset(mods_test_R, year == '2022'), aes(x = model, y = Rval_overall_noREs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('R (Testing dataset)') +
  ggtitle('Overall model, no random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
  
plot(p2)
  
  
p3 <- ggplot(subset(mods_test_R, year == '2022'), aes(x = model, y = Rval_cond_REs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('R (Testing dataset)') +
  ggtitle('Conditional model, with random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
  
plot(p3)

p4 <- ggplot(subset(mods_test_R, year == '2022'), aes(x = model, y = Rval_cond_noREs))+
  geom_hline(yintercept = 0) +
  geom_violin(draw_quantiles = c(0.5)) +
  geom_jitter(height = 0, width = 0.1, size = 3, aes(color = garden, shape = year)) +
  scale_color_manual(values = gard_cols, name = 'Garden') +
  scale_shape_manual(values = c(16,1), name = 'Year') +
  ylim(-1, 1.2) +
  ylab('R (Testing dataset)') +
  ggtitle('Conditional model, no random effects') +
  stat_compare_means(comparisons = my_comparisons, label = 'p.format', method = 'wilcox.test', paired = T)
plot(p4)


wrap_plots(p1, p2, p3, p4)

#ggsave(filename = paste('results/model_prediction/Rvalues_predictWithoutOneGarden_compareClimateGeneticsModels2022_', garden_clim_colname, '_vs_', pheno_colname, '.png', sep = ''), device = 'png', height = 14, width = 16)

########

# run linear model to test significance
anova(lm(Rval_overall_REs ~ model * year, data = mods_test_R))
anova(lm(Rval_overall_noREs ~ model*year, data = mods_test_R))

pairwise.t.test(mods_test_R$Rval_overall_noREs, mods_test_R$model)
anova(lm(Rval_cond_REs ~ model, data = mods_test_R))
anova(lm(Rval_cond_noREs ~ model, data = mods_test_R))
pairwise.t.test(mods_test_R$Rval_cond_noREs, mods_test_R$model)

```

# Predicting phenotypes based on model

## Function for predicting the response to climate

```{r predict_functions}

# predict each genotype's response to a vector of climate (MCMT) values
# garden_clims is a vector
# can calculate standard error of predictions using se.fit = TRUE - this takes longer to run

predict_genotype <- function(model,
                             type = 'response', # can also be conditional or zprob
                             effects,
                             home_clim,
                             garden_clims,
                             se.fit = FALSE){
  
  # vector to save predictions and their standard errors
  pred_height <- vector()
  pred_se <- vector()
  
  
  # loop through garden climates
  for(g in 1:length(garden_clims)){
    
    home <- home_clim
    garden <- garden_clims[g]
    
    # make newdata to give to predict()
    new_data <- as.data.frame(cbind(home, garden, home^2, garden^2))
    colnames(new_data) <- c('home_clim', 'garden_clim', 'home_clim_2', 'garden_clim_2')
    # add other random or fixed effects
    for(n in 1:length(effects)){
      new_data[names(effects[n])] <- effects[n]
    }
    
    # first have to scale new_data to match the scaled data used in model
    for(v in 1:length(new_data)){
      var <- names(new_data)[v]
      
      # only scale if var exists in scaling_factor - random effects like genotype are not in scaling factor
      # IF VARIABLES ARE NAMED DIFFERENTLY HERE IN effects AND scaling_factor IT WILL SKIP THEM
      if(var %in% names(scaling_factor)){
        new_data[var] <- new_data[var]*scaling_factor[[var]]
      } else {
        warning(var, ' does not have a scaling factor, skipping')
      }
      
      
    }
    
    # predict response for this new data
    pred <- predict(mod,
                    type = type,
                    newdata = new_data,
                    re.form = NA,
                    allow.new.levels=TRUE,
                    se.fit = se.fit)
    
    if(se.fit == TRUE) {
      fit <- pred$fit
      se <- pred$se.fit
    } else {
      fit <- pred
      se <- NA
    }
    
    # back transform from log if we're predicting height
    # zprob is probability of a zero, so don't transform
    if(type %in% c('response', 'conditional')){
      fit <- exp(fit) - 1
      # if we calculated standard error, transform that too
      if(se.fit == TRUE){
        se <- exp(se) - 1
      }
    }
    pred_height[g] <- fit
    pred_se[g] <- se
  }
  # save
  pred_height_df <- data.frame(garden_clim = garden_clims, predicted_height = pred_height, pred_se = pred_se)
  
  return(pred_height_df)
  
  # for troubleshooting - also return input data
  # return(list(predictions = pred_height_df,
  #             newdata = new_data))
}


```

## Setup

```{r predict_by_genotype_setup}

# get info for each genotype for use in predictions
geno_info <- as.data.frame(matrix(nrow= length(unique(dat$Genotype)), ncol = 4))
colnames(geno_info) <- c('pc1', 'pc2', 'pc3', 'home_clim')

rownames(geno_info) <- unique(dat$Genotype)

for(n in 1:nrow(geno_info)){
  
  geno <- rownames(geno_info)[n]
  
  # get first row with info for this genotype
  geno_row <- dat[match(geno, dat$Genotype),]
  
  geno_info$pc1[n] <- geno_row$genetic_PC1
  geno_info$pc2[n] <- geno_row$genetic_PC2
  geno_info$pc3[n] <- geno_row$genetic_PC3
  geno_info$home_clim[n] <- geno_row[,home_clim_colname]
  geno_info$Pt[n] <- geno_row$Pt
  
}

# remove the ones with missing data
geno_info <- geno_info[complete.cases(geno_info),]

# how many predictions to make for each genotype? (higher = smoother curves, but takes longer to run)
n_clims <- 100

############################################################# 
#overall prediction (incorporating both height and mortality)


# range of climate to predict for
# use range of both gardens and home climates
pred_clim_min <- range(c(garden_clim, home_clim), na.rm = T)[1]
pred_clim_max <- range(c(garden_clim, home_clim), na.rm = T)[2]

```


## Predict response to climate by genotype

```{r predict_by_genotype, warning = FALSE}

###############################################

# predictions for overall model (incorporating both height and mortality probability - makes predictions for height in cm)

# loop through genotypes and predict genotype-specific responses across the climate gradient (MCMT)
preds <- list()

par(mfrow = c(3,4))

for(n in 1:nrow(geno_info)){
  
  geno <- rownames(geno_info)[n]
  
  g.pc1 <- geno_info$pc1[n]
  g.pc2 <- geno_info$pc2[n]
  g.pc3 <- geno_info$pc3[n]
  g.home_clim <- geno_info$home_clim[n]
  
  
  
  preds[[n]] <- predict_genotype(mod, 
                                 effects = list(pc1 = g.pc1, 
                                                pc2 = g.pc2,
                                                pc3 = g.pc3,
                                                genotype = geno), 
                                 home_clim = g.home_clim, 
                                 garden_clims = seq(pred_clim_min, pred_clim_max, length.out = n_clims),
                                 se.fit = FALSE)
  
  # plot each genotype when done predicting
  # plot(preds[[n]]$garden_clim, preds[[n]]$predicted_height, main = rownames(geno_info)[n])
  # # lines for SE
  # lines(preds[[n]]$garden_clim, preds[[n]]$predicted_height+preds[[n]]$pred_se, col = 'blue')
  # lines(preds[[n]]$garden_clim, preds[[n]]$predicted_height-preds[[n]]$pred_se, col = 'blue')
  # 
  cat(paste('done with', n, '\n'))
  
}

names(preds) <- paste('genotype', rownames(geno_info), sep = '_')

  
######################
# plot all genotypes on one graph

colf <- colorRamp2(breaks = c(1, 0.5, 0), colors = c('#BCEE68', '#333333', '#1C86EE'))

#png(file = paste('results/model_prediction/predicted_height_by_genotype_overall_response_', clim_label, '.png', sep = ''), height = 8, width = 10, res = 300, units = 'in')
par(mfrow = c(1,1), cex.lab = 1.5)
plot(0,0,
     xlim = c(pred_clim_min, pred_clim_max), 
     ylim = c(0, ceiling(max(sapply(1:length(preds), function(n) max(preds[[n]]$predicted_height))))), 
     type = 'n', 
     xlab = clim_label, 
     ylab = paste(pheno_label, '(cm)'))

for(n in 1:length(preds)){
  
  lines(preds[[n]]$garden_clim, 
        preds[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}


# plot range of home climates
#abline(v = range(geno_info$home_clim, na.rm = T), lty = 2)
points(x = geno_info$home_clim, y = rep(0, nrow(geno_info)))
# plot range of garden climates
# abline(v = range(garden_clim, na.rm = T), lty = 1)
points(x = unique(garden_clim), y = rep(0, length(unique(garden_clim))), pch = '|')
legend('topright', pch =c('', '|'), legend = c('Home Climate', 'Garden Climate'))

  
#dev.off()

# plot with transfer distance instead of garden climate

# calculate x axis limits
xmin <- pred_clim_min - max(dat[,home_clim_colname])
xmax <- pred_clim_max - min(dat[,home_clim_colname])

#png(file = paste('results/model_prediction/predicted_height_by_genotype_transfer_dist_overall_response_', clim_label, '.png', sep = ''), height = 8, width = 10, res = 300, units = 'in')
par(mfrow = c(1,1), cex.lab = 1.5)
plot(0,0,
     xlim = c(xmin, xmax), 
     ylim = c(0, ceiling(max(sapply(1:length(preds), function(n) max(preds[[n]]$predicted_height))))), 
     type = 'n', 
     xlab = paste(clim_label, 'Transfer Distance'), 
     ylab = paste(pheno_label, '(cm)'))

abline(v = 0, lty = 1, col = 'grey40')

for(n in 1:length(preds)){
  
  lines(preds[[n]]$garden_clim - geno_info$home_clim[n], 
        preds[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

#dev.off()


# plot each individual separately, with real data points

# set shape for each year
shapes <- c(19, 17, 15)
cols.yr <- c('black', 'blue', 'red')

#pdf(file = paste('results/model_prediction/model_and_data_by_genotype_overall_', garden_clim_colname, '_', pheno_colname, '.pdf', sep = ''), height = 10, width = 12)
par(mfrow = c(3,4))
for(n in 1:length(preds)){

  # get genotype
  geno <- gsub('genotype_', '', names(preds)[n])
  
  # get genotype info
  sub <- df[df$genotype == geno,]
  
  # get R and p-value (calculated in model_plots chunk)
  Rval <- r_vs_pt[geno, 'cor_R']
  pval <- r_vs_pt[geno, 'cor_pval']
  
  plot(sub$garden_clim, sub$pheno, 
       pch = shapes[as.factor(sub$year)], 
       col = cols.yr[as.factor(sub$year)],
       #col = dat[dat$Genotype == geno, 'color_Pt'][1],
       xlab = paste('Garden', clim_label),
       ylab = pheno_label)
  title(main = paste('Genotype ', geno, '\n', 'R = ', Rval, '  |  p = ', pval, sep = ''), adj = 0)
  
  lines(preds[[n]]$garden_clim, 
        preds[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
  # SE lines
  #   lines(preds[[n]]$garden_clim, preds[[n]]$predicted_height+preds[[n]]$pred_se, col = 'grey')
  # lines(preds[[n]]$garden_clim, preds[[n]]$predicted_height-preds[[n]]$pred_se, col = 'grey')
  
  # line for home climate
  abline(v = sub$home_clim[1], lty = 2, col = 'grey')
  

  
}

# dev.off()

# does each genotype outcompete others at its home site?

#pdf(file = paste('results/model_prediction/model_compare_genotype_performance_at_home_', garden_clim_colname, '_', pheno_colname, '.pdf', sep = ''), height = 10, width = 12)
#png(file = paste('results/model_prediction/model_compare_genotype_performance_at_home_onePage', garden_clim_colname, '_', pheno_colname, '.png', sep = ''), height = 24, width = 16, units = 'in', res = 600)

#par(mfrow = c(9,5))
par(mfrow = c(3,4))
for(n in 1:length(preds)){
  
  # get genotype
  geno <- gsub('genotype_', '', names(preds)[n])
  
  # get genotype info
  sub <- df[df$genotype == geno,]
  
  plot(sub$garden_clim, sub$pheno, 
       type = 'n',
       #pch = shapes[as.factor(sub$year)], 
       #col = cols.yr[as.factor(sub$year)],
       #col = dat[dat$Genotype == geno, 'color_Pt'][1],
       xlab = paste('Garden', clim_label),
       ylab = pheno_label,
       xlim = c(pred_clim_min, pred_clim_max),
       ylim = c(0, ceiling(max(sapply(1:length(preds), function(n) max(preds[[n]]$predicted_height))))))
  title(main = paste('Genotype', geno), adj = 0)
  
  # line for home climate
  abline(v = sub$home_clim[1], lty = 2, lwd = 2, col = 'black')
  
  # add other genotypes
  for(g in 1:length(preds)){
    
    lines(preds[[g]]$garden_clim, 
          preds[[g]]$predicted_height, 
          col = colf(geno_info[g, 'Pt']), 
          lwd = 1,
          lty = 2)
  }
  
  # line for this genotype
  lines(preds[[n]]$garden_clim, 
        preds[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 3)
  
  
}



#dev.off()

###############################################

# predictions for conditional model (just height)

# loop through and predict
preds.cond <- list()

par(mfrow = c(3,4))
for(n in 1:nrow(geno_info)){
  
  geno <- rownames(geno_info)[n]
  
  g.pc1 <- geno_info$pc1[n]
  g.pc2 <- geno_info$pc2[n]
  g.pc3 <- geno_info$pc3[n]
  g.home_clim <- geno_info$home_clim[n]
  
  
  preds.cond[[n]] <- predict_genotype(mod, 
                                      type = 'conditional', 
                                      effects = list(pc1 = g.pc1, 
                                                     pc2 = g.pc2, 
                                                     pc3 = g.pc3, 
                                                     genotype = geno), 
                                      home_clim = g.home_clim, 
                                      garden_clims = seq(pred_clim_min, pred_clim_max, length.out = n_clims,
                                                         se.fit = FALSE))
  
  #plot(preds.cond[[n]]$garden_clim, preds.cond[[n]]$predicted_height, main = rownames(geno_info)[n])
  
  #cat(paste('done with', n, '\n'))
  
}

names(preds.cond) <- paste('genotype', rownames(geno_info), sep = '_')

# plot conditional model


#png(file = paste('results/model_prediction/predicted_height_by_genotype_conditional_model_', clim_label, '.png', sep = ''), height = 8, width = 10, res = 300, units = 'in')
par(mfrow = c(1,1), cex.lab = 1.5)

plot(0,0, xlim = c(pred_clim_min, pred_clim_max), 
     ylim = c(0, ceiling(max(sapply(1:length(preds.cond), function(n) max(preds.cond[[n]]$predicted_height))))), 
     type = 'n', 
     xlab = clim_label,
     ylab = paste(pheno_label, '(cm)'))

for(n in 1:length(preds.cond)){
  
  lines(preds.cond[[n]]$garden_clim, 
        preds.cond[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

# plot range of home climates
#abline(v = range(geno_info$home_clim, na.rm = T), lty = 2)
points(x = geno_info$home_clim, y = rep(0, nrow(geno_info)))
# plot range of garden climates
#abline(v = range(garden_clim, na.rm = T), lty = 1)
points(x = unique(garden_clim), y = rep( 0, length(unique(garden_clim))), pch = '|')
legend('topright', pch =c('', '|'), legend = c('Home Climate', 'Garden Climate'))

 
#dev.off()

# plot with transfer distance instead of garden climate

# calculate x axis limits
xmin <- pred_clim_min - max(dat[,home_clim_colname])
xmax <- pred_clim_max - min(dat[,home_clim_colname])

#png(file = paste('results/model_prediction/predicted_height_by_genotype_transfer_dist_conditional_model_', clim_label, '.png', sep = ''), height = 8, width = 10, res = 300, units = 'in')
par(mfrow = c(1,1), cex.lab = 1.5)
plot(0,0,
     xlim = c(xmin, xmax), 
     ylim = c(0,ceiling(max(sapply(1:length(preds.cond), function(n) max(preds.cond[[n]]$predicted_height))))), 
     type = 'n', 
     xlab = paste(clim_label, 'Transfer Distance'), 
     ylab = paste(pheno_label, '(cm)'))

abline(v = 0, col = 'grey40')

for(n in 1:length(preds.cond)){
  
  lines(preds[[n]]$garden_clim - geno_info$home_clim[n], 
        preds.cond[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

#dev.off()


# each genotype individually

#pdf(file = paste('results/model_prediction/model_and_data_by_genotype_conditional_', garden_clim_colname, '_', pheno_colname, '.pdf', sep = ''), height = 10, width = 12)
par(mfrow = c(3,4))
for(n in 1:length(preds)){

  # get genotype
  geno <- gsub('genotype_', '', names(preds)[n])
  
  # get genotype info
  sub <- df[df$genotype == geno,]
  
  # get R and p-value (calculated in model_plots chunk)
  # Rval <- r_vs_pt[geno, 'cor_R']
  # pval <- r_vs_pt[geno, 'cor_pval']
  
  plot(sub$garden_clim, sub$pheno, 
       #pch = 16, 
       col = cols.yr[as.factor(sub$year)],
       pch = shapes[as.factor(sub$year)],
       #col = sub$color_Pt,
       xlab = paste('Garden', clim_label),
       ylab = pheno_label,
       main = paste('Genotype', geno))
  # haven't calculated R and p values for the conditional model yet, so don't plot them
  #title(main = paste('Genotype ', geno, '\n', 'R = ', Rval, '  |  p = ', pval, sep = ''), adj = 0)
  
  lines(preds[[n]]$garden_clim, preds[[n]]$predicted_height, col = colf(geno_info[n, 'Pt']), lwd = 2, lty = 2)
  lines(preds.cond[[n]]$garden_clim, preds.cond[[n]]$predicted_height, col = colf(geno_info[n, 'Pt']), lwd = 2)
  
  # line for home climate
  abline(v = sub$home_clim[1], lty = 2, col = 'grey')
  
}

#dev.off()


#############################################
# predict for zero inflated model (mortality)

# loop through and predict
preds.zi <- list()

par(mfrow = c(3,4))
for(n in 1:nrow(geno_info)){
  
  geno <- rownames(geno_info)[n]
  
  g.pc1 <- geno_info$pc1[n]
  g.pc2 <- geno_info$pc2[n]
  g.pc3 <- geno_info$pc3[n]
  g.home_clim <- geno_info$home_clim[n]
  
  preds.zi[[n]] <- predict_genotype(mod, 
                                    type = 'zprob', 
                                    effects = list(pc1 = g.pc1, 
                                                   pc2 = g.pc2, 
                                                   pc3 = g.pc3, 
                                                   genotype = geno), 
                                    home_clim = g.home_clim, 
                                    garden_clims = seq(pred_clim_min, pred_clim_max, length.out = n_clims),
                                    se.fit = FALSE)
  
  #plot(preds.zi[[n]]$garden_clim, preds.zi[[n]]$predicted_height, main = rownames(geno_info)[n], ylim = c(0,1))
  
  #cat(paste('done with', n, '\n'))
  
}

names(preds.zi) <- paste('genotype', rownames(geno_info), sep = '_')

# plot zero-inflated model


#png(file = paste('results/model_prediction/predicted_height_by_genotype_zero-inflated_', clim_label, '.png', sep = ''), height = 8, width = 10, res = 300, units = 'in')
par(mfrow = c(1,1), cex.lab = 1.5)
plot(0,0,
     xlim = c(pred_clim_min, pred_clim_max), 
     ylim = c(0,1), 
     type = 'n', 
     xlab = clim_label, 
     ylab = 'Probablility of Mortality')

for(n in 1:length(preds.zi)){
  
  lines(preds.zi[[n]]$garden_clim, 
        preds.zi[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

# plot range of home climates
#abline(v = range(geno_info$home_clim, na.rm = T), lty = 2)
points(x = geno_info$home_clim, y = rep(0, nrow(geno_info)))
# plot range of garden climates
#abline(v = range(garden_clim, na.rm = T), lty = 1)
points(x = unique(garden_clim), y = rep( 0, length(unique(garden_clim))), pch = '|')

legend('topright', pch =c('', '|'), legend = c('Home Climate', 'Garden Climate'))

 
#dev.off()

# plot with transfer distance instead of garden climate

# calculate x axis limits
xmin <- pred_clim_min - max(dat[,home_clim_colname])
xmax <- pred_clim_max - min(dat[,home_clim_colname])

#png(file = paste('results/model_prediction/predicted_height_by_genotype_transfer_dist_zi_response_', clim_label, '.png', sep = ''), height = 8, width = 10, res = 300, units = 'in')
par(mfrow = c(1,1), cex.lab = 1.5)
plot(0,0,
     xlim = c(xmin, xmax), 
     ylim = c(0,1), 
     type = 'n', 
     xlab = paste(clim_label, 'Transfer Distance'), 
     ylab = 'Probability of Mortality')

abline(v = 0, col = 'grey40')

for(n in 1:length(preds.zi)){
  
  lines(preds.zi[[n]]$garden_clim - geno_info$home_clim[n], 
        preds.zi[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

#dev.off()


#################################
# calculate optimal climate
geno_info$optimal_clim <- NA
for(n in 1:length(preds)){
  
  tmp <- preds[[n]]
  geno_info$optimal_clim[n] <- tmp[which.max(tmp$predicted_height), 'garden_clim']
  
}

plot(geno_info$home_clim, geno_info$optimal_clim, col = colf(geno_info$Pt), pch = 16, xlab = 'home climate', ylab = 'predicted optimal climate', cex = 1.5)

plot(geno_info$home_clim, geno_info$optimal_clim, type = 'n', xlab = 'home climate', ylab = 'predicted optimal climate', cex = 1.5)
text(geno_info$home_clim, geno_info$optimal_clim, labels = rownames(geno_info), col = colf(geno_info$Pt))
abline(0,1,lty = 2)

# save

#predictions <- list(overall = preds, conditional = preds.cond, zi = preds.zi, genotypes = geno_info)
#save(predictions, file = paste('results/model_prediction/predictedGrowth_acrossClimates_byGenotype_', garden_clim_colname, '.Rdata', sep = ''))

```

```{r panel_plot}


# to skip running the predictions, which take a long time, can load output file from the above chunk
#load(paste('results/model_prediction/predictedGrowth_acrossClimates_byGenotype_', garden_clim_colname, '.Rdata', sep = ''))
# preds <- predictions$overall
# preds.cond <- predictions$conditional
# preds.zi <- predictions$zi

colf <- colorRamp2(breaks = c(1, 0.5, 0), colors = c('#BCEE68', '#333333', '#1C86EE'))

#png(file = paste('results/model_prediction/predicted_height_by_genotype_panel_', garden_clim_colname, sep = ''), height = 7, width = 10, res = 300, units = 'in')

par(mfrow = c(2,3), cex.lab = 1.5, cex.main = 1.5)

# overall
plot(0,0,
     xlim = c(pred_clim_min, pred_clim_max), 
     ylim = c(0, ceiling(max(sapply(1:length(preds.cond), function(n) max(preds.cond[[n]]$predicted_height))))), 
     type = 'n', 
     xlab = paste(clim_label, '(C)'), 
     ylab = paste(pheno_label, '(cm)'))
title('A. Overall model', adj = 0)

for(n in 1:length(preds)){
  
  lines(preds[[n]]$garden_clim, 
        preds[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}


# plot range of home climates
#abline(v = range(geno_info$home_clim, na.rm = T), lty = 2)
points(x = geno_info$home_clim, y = rep(0, nrow(geno_info)), pch = '|')
# plot range of garden climates
# abline(v = range(garden_clim, na.rm = T), lty = 1)
points(x = unique(garden_clim), y = rep(0, length(unique(garden_clim))), pch = 1)
legend('topright', pch =c('|', ''), legend = c('Home Climate', 'Garden Climate'))

  


# conditional
plot(0,0, xlim = c(pred_clim_min, pred_clim_max), 
     ylim = c(0, ceiling(max(sapply(1:length(preds.cond), function(n) max(preds.cond[[n]]$predicted_height))))),
     type = 'n', 
     xlab = paste(clim_label, '(C)'),
     ylab = paste(pheno_label, '(cm)'))
title('B. Conditional model (growth)', adj = 0)

for(n in 1:length(preds.cond)){
  
  lines(preds.cond[[n]]$garden_clim, 
        preds.cond[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

# plot range of home climates
#abline(v = range(geno_info$home_clim, na.rm = T), lty = 2)
points(x = geno_info$home_clim, y = rep(0, nrow(geno_info)), pch = '|')
# plot range of garden climates
# abline(v = range(garden_clim, na.rm = T), lty = 1)
points(x = unique(garden_clim), y = rep(0, length(unique(garden_clim))), pch = 1)

#legend('topright', pch =c(1, 16), legend = c('Home Climate', 'Garden Climate'))


# zi
plot(0,0,
     xlim = c(pred_clim_min, pred_clim_max), 
     ylim = c(0,1), 
     type = 'n', 
     xlab = paste(clim_label, '(C)'), 
     ylab = 'Probablility of Mortality')
title('C. Zero-inflated Model (mortality)', adj = 0)

for(n in 1:length(preds.cond)){
  
  lines(preds.zi[[n]]$garden_clim, 
        preds.zi[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

# plot range of home climates
#abline(v = range(geno_info$home_clim, na.rm = T), lty = 2)
points(x = geno_info$home_clim, y = rep(0, nrow(geno_info)), pch = '|')
# plot range of garden climates
# abline(v = range(garden_clim, na.rm = T), lty = 1)
points(x = unique(garden_clim), y = rep(0, length(unique(garden_clim))), pch = 1)

#legend('topright', pch =c(1, 16), legend = c('Home Climate', 'Garden Climate'))

#################

# plot with transfer distance instead of garden climate

# overall
# calculate x axis limits
xmin <- pred_clim_min - max(dat[,home_clim_colname])
xmax <- pred_clim_max - min(dat[,home_clim_colname])

plot(0,0,
     xlim = c(xmin, xmax), 
     ylim = c(0,ceiling(max(sapply(1:length(preds.cond), function(n) max(preds.cond[[n]]$predicted_height))))), 
     type = 'n', 
     xlab = paste(clim_label, 'Transfer Distance (C)'), 
     ylab = paste(pheno_label, '(cm)'))
title('D.', adj = 0)

abline(v = 0, lty = 1, col = 'grey40')

for(n in 1:length(preds)){
  
  lines(preds[[n]]$garden_clim - geno_info$home_clim[n], 
        preds[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

# plot with transfer distance instead of garden climate

# calculate x axis limits
xmin <- pred_clim_min - max(dat[,home_clim_colname])
xmax <- pred_clim_max - min(dat[,home_clim_colname])


plot(0,0,
     xlim = c(xmin, xmax), 
     ylim = c(0, ceiling(max(sapply(1:length(preds.cond), function(n) max(preds.cond[[n]]$predicted_height))))), 
     type = 'n', 
     xlab = paste(clim_label, 'Transfer Distance (C)'), 
     ylab = paste(pheno_label, '(cm)'))

title('E.', adj = 0)

abline(v = 0, col = 'grey40')

for(n in 1:length(preds.cond)){
  
  lines(preds[[n]]$garden_clim - geno_info$home_clim[n], 
        preds.cond[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

# plot with transfer distance instead of garden climate

# calculate x axis limits
xmin <- pred_clim_min - max(dat[,home_clim_colname])
xmax <- pred_clim_max - min(dat[,home_clim_colname])


plot(0,0,
     xlim = c(xmin, xmax), 
     ylim = c(0,1), 
     type = 'n', 
     xlab = paste(clim_label, 'Transfer Distance (C)'), 
     ylab = 'Probability of Mortality')

title('F.', adj = 0)

abline(v = 0, col = 'grey40')

for(n in 1:length(preds.zi)){
  
  lines(preds.zi[[n]]$garden_clim - geno_info$home_clim[n], 
        preds.zi[[n]]$predicted_height, 
        col = colf(geno_info[n, 'Pt']), 
        lwd = 2)
  
}

#dev.off()

```
